---
title: "MSstats PSM-Level Proteomics Analysis"
lightbox: true
subtitle: "Reusable analysis driven by PSM and annotation files"
author:
  - Kevin Cormier
date: today
title-block-banner: "#27445C"
format:
  html:
    toc: true
    toc-location: left
    toc-active-border: red
    page-layout: full
    theme:
      light: [cosmo, assets/offline-fonts.scss]
      dark: [darkly, assets/offline-fonts.scss]
    html-math-method: katex
    code-fold: true
    highlight-style: tango
    toc_float: true
    number-sections: false
    embed-resources: true
    smooth-scroll: true
    fig-align: 'center'
    fig-width: 10
    fig-height: 10
css: assets/style.css
editor: source
execute:
  cache: false
editor_options:
  chunk_output_type: console
params:
  request: "msstats-report"
  report_label: "MSstats analysis"
  psm_file: null
  psm_sheet: "PSMs"
  annotation_file: null
  annotation_sheet: 1
  condition_columns: null
  condition_template: null
  condition_separator: "_"
  output_dir: "results/msstats-report"
  figures_dir: "figures/msstats-report"
  contrast_strategy: "all_pairwise"
  contrasts: null
  weighted_contrasts: null
  contrast_file: null
  excluded_runs: null
  organism_package: "org.Hs.eg.db"
  run_gsea: true
  max_psm_rows: null
  filter_confidence: "High"
  remove_contaminants: true
  remove_shared_peptides: true
  remove_single_peptide_proteins: false
  remove_few_measurements: false
  normalization: "equalizeMedians"
  summary_method: "TMP"
  mb_impute: true
  number_of_cores: 1
  volcano_p_cutoff: 0.05
  volcano_fc_cutoff: 1
  volcano_top_labels: 10
  volcano_split_axis_quantile: 0.95
  volcano_split_axis_min_ratio: 1.8
  volcano_split_axis_scales: 0.35
  qc_max_features: 5000
  qc_top_cv_proteins: 25
  power_alpha: 0.05
  power_target: 0.8
  power_effect_log2fc: 1
  power_top_rows: 200
  gsea_pvalue_cutoff: 0.05
  gsea_min_gene_set_size: 100
  gsea_max_gene_set_size: 500
  gsea_dot_show_categories: 10
  gsea_tree_show_categories: 20
  gsea_tree_label_wrap: 40
---

```{r, include=FALSE}
if (.Platform$OS.type == "windows") {
  options(download.file.method = "wininet")
}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = FALSE, fig.align = 'center',
                      fig.width = 10, fig.height = 10,
                      comment = NA)
```

```{r setup}
required_packages <- c(
  "MSstats",
  "data.table",
  "readxl",
  "dplyr",
  "tidyr",
  "stringr",
  "purrr",
  "ggplot2",
  "ggrepel",
  "scales",
  "EnhancedVolcano",
  "ggbreak",
  "readr",
  "DT",
  "knitr",
  "tibble",
  "clusterProfiler",
  "AnnotationDbi",
  params$organism_package,
  "enrichplot",
  "htmltools"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop("Install required package(s): ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

suppressPackageStartupMessages({
  library(MSstats)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(EnhancedVolcano)
  library(ggbreak)
  library(readr)
  library(DT)
  library(knitr)
  library(tibble)
})

current_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
report_dir <- if (!is.null(current_input) && nzchar(current_input)) {
  dirname(normalizePath(current_input, winslash = "/", mustWork = TRUE))
} else {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}

required_path <- function(value, parameter_name) {
  if (is.null(value) || length(value) != 1 || is.na(value) || !nzchar(value)) {
    stop("Set the ", parameter_name, " parameter before rendering.", call. = FALSE)
  }
  value
}

resolve_path <- function(path) {
  path <- required_path(path, "file path")
  if (grepl("^(/|[A-Za-z]:[/\\\\])", path)) path else file.path(report_dir, path)
}

psm_file <- normalizePath(
  resolve_path(required_path(params$psm_file, "psm_file")),
  winslash = "/",
  mustWork = TRUE
)
annotation_file <- normalizePath(
  resolve_path(required_path(params$annotation_file, "annotation_file")),
  winslash = "/",
  mustWork = TRUE
)
output_dir <- resolve_path(params$output_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
read_psm_n_max <- if (is.null(params$max_psm_rows)) Inf else params$max_psm_rows
source(file.path(report_dir, "R", "pathway-analysis-functions.R"), local = knitr::knit_global())
source(file.path(report_dir, "R", "contrast-functions.R"), local = knitr::knit_global())
source(file.path(report_dir, "R", "input-functions.R"), local = knitr::knit_global())
organism_db <- getExportedValue(params$organism_package, params$organism_package)

figure_dir <- resolve_path(params$figures_dir)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  fig.path = file.path(params$figures_dir, paste0(params$request, "-")),
  dev = c("png", "pdf"),
  dpi = 150
)

set.seed(20250707)
```

## Methods

The Protein Spectral Match (PSM-level) reporter-ion intensities were imported from the Proteome Discoverer
export and linked to experimental condition, biological replicate, and raw-file
metadata using the annotation workbook. Quantitative rows with missing or
non-positive abundance were removed; when available, only `r
params$filter_confidence`-confidence PSMs were retained and contaminant-flagged
rows were excluded. Peptides shared among proteins were removed during
conversion to MSstats, while proteins represented by a single peptide were
retained.

Proteome Discoverer fields were converted with
`MSstats::PDtoMSstatsFormat()`, resolving duplicate feature measurements by
their maximum intensity. Intensities were log2 transformed, normalized by
`r params$normalization`, summarized from peptide features to proteins with the
`r params$summary_method` method, and missing values were handled using
MSstats model-based imputation. Quality assessment included run-level feature
counts and intensity distributions, missingness, coefficients of variation,
Pearson correlations, and principal component analysis.

Condition names were read from the matched annotation rows. Comparisons were
either supplied in the parameter file or created automatically from those
conditions. MSstats supplied log2 fold changes, standard errors, test
statistics, degrees of freedom, raw p-values, and Benjamini-Hochberg adjusted
p-values. Differential expression was defined as an adjusted p-value below `r
params$volcano_p_cutoff` and an absolute log2 fold change of at least `r
params$volcano_fc_cutoff`.

When enabled, GO Biological Process gene set enrichment analysis (GSEA) mapped
UniProt accessions with `r params$organism_package` and ranked them by signed
statistical evidence. Gene sets containing `r params$gsea_min_gene_set_size` to
`r params$gsea_max_gene_set_size` genes were tested with clusterProfiler.
Power diagnostics used a two-sided alpha of `r params$power_alpha`, target
power of `r params$power_target`, and target absolute effect of `r
params$power_effect_log2fc` log2 units.

## Analysis Inputs

This report analyzes `r params$report_label` using:

- PSM export: `r params$psm_file`
- Annotation file: `r params$annotation_file`

The Proteome Discoverer export uses `File ID`, `Quan Value`, `Master Protein Accessions`, and `Annotated Sequence`; these are standardized for `MSstats::PDtoMSstatsFormat()` below.

```{r load-data}
psm_raw <- read_input_table(
  psm_file,
  sheet = params$psm_sheet,
  n_max = read_psm_n_max
)

annotation_raw <- read_input_table(
  annotation_file,
  sheet = params$annotation_sheet
)

required_psm_cols <- c(
  "File ID",
  "Annotated Sequence",
  "Charge",
  "Master Protein Accessions",
  "Quan Value"
)
condition_columns <- as.character(params$condition_columns)
condition_columns <- condition_columns[!is.na(condition_columns) & nzchar(condition_columns)]
required_annotation_cols <- c("File Name", "File ID", "BioReplicate")
if (length(condition_columns)) {
  required_annotation_cols <- c(required_annotation_cols, condition_columns)
} else {
  required_annotation_cols <- c(required_annotation_cols, "Condition")
}

missing_psm_cols <- setdiff(required_psm_cols, names(psm_raw))
missing_annotation_cols <- setdiff(required_annotation_cols, names(annotation_raw))
if (length(missing_psm_cols) > 0) {
  stop("Missing required PSM column(s): ", paste(missing_psm_cols, collapse = "; "), call. = FALSE)
}
if (length(missing_annotation_cols) > 0) {
  stop("Missing required annotation column(s): ", paste(missing_annotation_cols, collapse = "; "), call. = FALSE)
}

annotation_raw$Condition <- make_condition(
  annotation_raw,
  condition_columns = condition_columns,
  condition_template = params$condition_template,
  condition_separator = params$condition_separator
)

psm <- psm_raw |>
  rename(SpectrumFile = `File ID`) |>
  mutate(
    SpectrumFile = as.character(SpectrumFile),
    `Quan Value` = suppressWarnings(as.numeric(`Quan Value`))
  )

annotation <- annotation_raw |>
  rename(Run = `File ID`) |>
  mutate(
    Run = as.character(Run),
    Condition = as.character(Condition),
    BioReplicate = as.character(BioReplicate),
    SampleName = tools::file_path_sans_ext(basename(as.character(`File Name`))),
    FigureLabel = .data$SampleName
  )

excluded_runs <- as.character(params$excluded_runs)
excluded_runs <- excluded_runs[!is.na(excluded_runs) & nzchar(excluded_runs)]
if (length(excluded_runs) > 0) {
  psm <- psm |>
    filter(!.data$SpectrumFile %in% excluded_runs)
}

missing_runs <- setdiff(unique(psm$SpectrumFile), annotation$Run)
if (length(missing_runs) > 0) {
  stop("Missing annotation for run(s): ", paste(missing_runs, collapse = ", "), call. = FALSE)
}

annotation <- annotation |>
  filter(.data$Run %in% unique(psm$SpectrumFile))

duplicate_annotation <- annotation |>
  count(.data$Run, name = "rows") |>
  filter(is.na(.data$Run) | !nzchar(.data$Run) | .data$rows > 1)
if (nrow(duplicate_annotation) > 0) {
  stop(
    "Each matched annotation File ID must be present exactly once. Check: ",
    paste(duplicate_annotation$Run, collapse = ", "),
    call. = FALSE
  )
}

if (any(is.na(annotation$Condition) | !nzchar(annotation$Condition))) {
  stop("Every matched annotation row must have a Condition value.", call. = FALSE)
}
```

```{r qc-helper-functions}
source(file.path(report_dir, "R", "qc-helper-functions.R"), local = knitr::knit_global())
```

## Sample Annotation

```{r sample-annotation}
annotation |>
  mutate(.bioreplicate_order = suppressWarnings(as.numeric(.data$BioReplicate))) |>
  arrange(
    .data$Condition,
    is.na(.data$.bioreplicate_order),
    .data$.bioreplicate_order,
    .data$BioReplicate,
    .data$`File Name`
  ) |>
  select(`File Name`, Condition, BioReplicate) |>
  knitr::kable(caption = "Sample annotation used in the MSstats analysis.")
```

## Raw Data Overview

```{r raw-summary}
psm_summary <- tibble(
  metric = c(
    "PSM rows",
    "Runs",
    "Conditions",
    "Protein accessions",
    "Annotated peptide sequences",
    "Non-missing quant values"
  ),
  value = c(
    nrow(psm),
    dplyr::n_distinct(psm$SpectrumFile),
    dplyr::n_distinct(annotation$Condition),
    dplyr::n_distinct(psm$`Master Protein Accessions`, na.rm = TRUE),
    dplyr::n_distinct(psm$`Annotated Sequence`, na.rm = TRUE),
    sum(!is.na(psm$`Quan Value`))
  )
)

knitr::kable(psm_summary, caption = "Summary of the raw PSM dataset.")
```

```{r annotation-summary}
annotation |>
  count(Condition, name = "runs") |>
  arrange(Condition) |>
  knitr::kable(caption = "Number of experimental runs by condition.")
```

```{r raw-psm-counts, fig.cap="PSM counts for each sample before filtering, colored by experimental condition."}
psm |>
  count(SpectrumFile, name = "psms") |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  ggplot(aes(x = reorder(FigureLabel, psms), y = psms, fill = Condition)) +
  geom_col() +
  coord_flip() +
  labs(x = "Sample", y = "PSM rows", title = "PSM rows per sample") +
  theme_bw()
```

## Raw Data QC

These checks look at the unfiltered PSM export. They are meant to catch run-level
dropout, low-intensity samples, contaminant burden, shared peptide burden, and
unexpected charge or modification patterns before any MSstats modeling.

```{r raw-missingness-zero-summary}
raw_intensity_qc <- psm |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  group_by(.data$SpectrumFile, .data$FigureLabel, .data$Condition) |>
  summarise(
    psm_rows = n(),
    missing_intensity = sum(is.na(.data$`Quan Value`)),
    non_positive_intensity = sum(!is.na(.data$`Quan Value`) & .data$`Quan Value` <= 0),
    positive_intensity = sum(!is.na(.data$`Quan Value`) & .data$`Quan Value` > 0),
    missing_or_non_positive_percent =
      100 * (.data$missing_intensity + .data$non_positive_intensity) / .data$psm_rows,
    .groups = "drop"
  ) |>
  arrange(.data$Condition, .data$SpectrumFile)

knitr::kable(
  raw_intensity_qc,
  digits = 2,
  caption = "Raw intensity completeness and non-positive values by sample."
)
```

```{r raw-missingness-zero-plot, fig.height=6, fig.cap="Raw PSM intensity-status counts per sample, separating positive, non-positive, and missing quantitative values."}
raw_intensity_qc |>
  tidyr::pivot_longer(
    cols = c("missing_intensity", "non_positive_intensity", "positive_intensity"),
    names_to = "status",
    values_to = "rows"
  ) |>
  mutate(
    status = factor(
      .data$status,
      levels = c("positive_intensity", "non_positive_intensity", "missing_intensity"),
      labels = c("Positive", "Non-positive", "Missing")
    )
  ) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$rows), y = .data$rows, fill = .data$status)) +
  geom_col() +
  coord_flip() +
  labs(x = "Sample", y = "PSM rows", fill = "Intensity status",
       title = "Raw PSM intensity status per sample") +
  theme_bw()
```

```{r raw-intensity-distribution, fig.height=7, fig.cap="Distribution of raw log2 PSM intensities for each sample, colored by condition."}
psm |>
  filter(!is.na(.data$`Quan Value`), .data$`Quan Value` > 0) |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  mutate(log2QuanValue = log2(.data$`Quan Value`)) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$log2QuanValue, median),
             y = .data$log2QuanValue, fill = .data$Condition)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.4) +
  coord_flip() +
  labs(x = "Sample", y = "log2(Quan Value)",
       title = "Raw intensity distribution per sample") +
  theme_bw()
```

```{r raw-intensity-density, fig.height=6, fig.cap="Kernel-density curves of raw log2 PSM intensities for each sample, colored by condition."}
psm |>
  filter(!is.na(.data$`Quan Value`), .data$`Quan Value` > 0) |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  mutate(log2QuanValue = log2(.data$`Quan Value`)) |>
  ggplot(aes(x = .data$log2QuanValue, color = .data$Condition, group = .data$SpectrumFile)) +
  geom_density(alpha = 0.45, linewidth = 0.5) +
  labs(x = "log2(Quan Value)", y = "Density",
       title = "Raw intensity density per run") +
  theme_bw()
```

```{r raw-contaminant-shared-single-summary}
contaminant_rows <- if ("Contaminant" %in% names(psm)) {
  sum(is_truthy_flag(psm$Contaminant))
} else {
  NA_integer_
}

shared_peptide_rows <- if ("# Proteins" %in% names(psm)) {
  sum(suppressWarnings(as.numeric(psm$`# Proteins`)) > 1, na.rm = TRUE)
} else {
  NA_integer_
}

single_peptide_proteins <- psm |>
  filter(!is.na(.data$`Master Protein Accessions`),
         !is.na(.data$`Annotated Sequence`)) |>
  distinct(.data$`Master Protein Accessions`, .data$`Annotated Sequence`) |>
  count(.data$`Master Protein Accessions`, name = "unique_peptides") |>
  summarise(n = sum(.data$unique_peptides == 1)) |>
  pull(.data$n)

tibble(
  metric = c(
    "Contaminant PSM rows",
    "Shared peptide PSM rows (# Proteins > 1)",
    "Proteins represented by one unique peptide"
  ),
  value = c(contaminant_rows, shared_peptide_rows, single_peptide_proteins)
) |>
  knitr::kable(
    caption = "Contaminant, shared-peptide, and single-peptide-protein summary."
  )
```

```{r raw-charge-distribution, fig.height=5, fig.cap="Raw PSM counts by precursor charge state."}
if ("Charge" %in% names(psm)) {
  psm |>
    mutate(Charge = as.character(.data$Charge)) |>
    count(.data$Charge, name = "psm_rows") |>
    ggplot(aes(x = .data$Charge, y = .data$psm_rows)) +
    geom_col(fill = "#4C72B0") +
    labs(x = "Charge", y = "PSM rows", title = "Charge-state distribution") +
    theme_bw()
} else {
  message("Charge-state plot skipped: Charge column was not found.")
}
```

```{r raw-modification-distribution, fig.height=5, fig.cap="Raw PSM counts for modified and unmodified peptide-spectrum matches."}
if ("Modifications" %in% names(psm)) {
  psm |>
    mutate(
      modification_status = if_else(has_modification(.data$Modifications),
                                    "Modified", "Unmodified")
    ) |>
    count(.data$modification_status, name = "psm_rows") |>
    ggplot(aes(x = .data$modification_status, y = .data$psm_rows,
               fill = .data$modification_status)) +
    geom_col(show.legend = FALSE) +
    labs(x = NULL, y = "PSM rows", title = "Modified vs unmodified PSM rows") +
    theme_bw()
} else {
  message("Modification plot skipped: Modifications column was not found.")
}
```

## Preprocessing for MSstats

The PSM table is filtered before conversion. By default this report keeps high-confidence PSMs, removes rows flagged as contaminants when that column is present, and removes non-positive or missing intensities.

```{r preprocess-for-msstats}
psm_filtered <- psm |>
  filter(!is.na(`Quan Value`), `Quan Value` > 0)

if (!is.null(params$filter_confidence) && "Confidence" %in% names(psm_filtered)) {
  psm_filtered <- psm_filtered |>
    filter(.data$Confidence == params$filter_confidence)
}

if (isTRUE(params$remove_contaminants) && "Contaminant" %in% names(psm_filtered)) {
  psm_filtered <- psm_filtered |>
    filter(is.na(.data$Contaminant) | .data$Contaminant == FALSE)
}

preprocess_summary <- tibble(
  step = c("Raw PSM rows", "Rows after filtering"),
  rows = c(nrow(psm), nrow(psm_filtered))
)

knitr::kable(
  preprocess_summary,
  caption = "PSM row counts before and after MSstats input filtering."
)
```

## Filtered Input QC

These summaries describe the data that will actually be submitted to
`MSstats::PDtoMSstatsFormat()`.

```{r filtered-feature-counts}
psm_filtered_features <- add_psm_feature_id(psm_filtered)

filtered_run_qc <- psm_filtered_features |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  group_by(.data$SpectrumFile, .data$FigureLabel, .data$Condition) |>
  summarise(
    psm_rows = n(),
    annotated_sequences = n_distinct(.data$`Annotated Sequence`, na.rm = TRUE),
    peptide_charge_features = n_distinct(.data$.qc_feature_id, na.rm = TRUE),
    protein_accessions = n_distinct(.data$`Master Protein Accessions`, na.rm = TRUE),
    median_log2_intensity = median(log2(.data$`Quan Value`), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(.data$Condition, .data$SpectrumFile)

knitr::kable(
  filtered_run_qc,
  digits = 2,
  caption = "Filtered MSstats input features and intensities by sample."
)
```

```{r filtered-feature-counts-plot, fig.height=7, fig.cap="Filtered PSM, annotated-sequence, peptide-charge-feature, and protein-accession counts for each sample."}
filtered_run_qc |>
  select(SpectrumFile, FigureLabel, Condition, psm_rows, annotated_sequences,
         peptide_charge_features, protein_accessions) |>
  tidyr::pivot_longer(
    cols = c("psm_rows", "annotated_sequences",
             "peptide_charge_features", "protein_accessions"),
    names_to = "metric",
    values_to = "count"
  ) |>
  mutate(
    metric = factor(
      .data$metric,
      levels = c("psm_rows", "annotated_sequences",
                 "peptide_charge_features", "protein_accessions"),
      labels = c("PSM rows", "Annotated sequences",
                 "Peptide-charge features", "Protein accessions")
    )
  ) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$count),
             y = .data$count, fill = .data$Condition)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free_x") +
  labs(x = "Sample", y = "Count",
       title = "Filtered PSM, peptide, feature, and protein counts per sample") +
  theme_bw()
```

```{r filtered-run-correlation, fig.height=8, fig.cap="Pearson correlations among filtered peptide-charge feature profiles; samples are labeled by condition and biological replicate."}
filtered_feature_matrix <- make_feature_run_matrix(
  psm_filtered_features,
  feature_col = ".qc_feature_id",
  run_col = "SpectrumFile",
  value_col = "Quan Value"
)

plot_correlation_heatmap(
  filtered_feature_matrix,
  annotation,
  title = "Filtered PSM feature correlation between runs"
)
```

```{r helper-functions}
convert_pd_to_msstats <- function(psm_table, annotation_table) {
  MSstats::PDtoMSstatsFormat(
    input = psm_table,
    annotation = annotation_table,
    useUniquePeptide = params$remove_shared_peptides,
    summaryforMultipleRows = max,
    removeFewMeasurements = params$remove_few_measurements,
    removeProtein_with1Peptide = params$remove_single_peptide_proteins,
    which.quantification = "Quan Value",
    which.proteinid = "Master Protein Accessions",
    which.sequence = "Annotated Sequence",
    use_log_file = FALSE,
    verbose = FALSE
  )
}

process_msstats <- function(msstats_input) {
  MSstats::dataProcess(
    raw = msstats_input,
    logTrans = 2,
    normalization = params$normalization,
    summaryMethod = params$summary_method,
    MBimpute = params$mb_impute,
    use_log_file = FALSE,
    verbose = FALSE,
    numberOfCores = params$number_of_cores
  )
}

make_contrast_matrix <- function(groups, contrast_table) {
  missing_groups <- setdiff(unique(c(contrast_table$numerator, contrast_table$denominator)), groups)
  if (length(missing_groups) > 0) {
    stop("Contrast group(s) not present in processed data: ", paste(missing_groups, collapse = ", "), call. = FALSE)
  }

  contrast_matrix <- matrix(
    0,
    nrow = nrow(contrast_table),
    ncol = length(groups),
    dimnames = list(contrast_table$label, groups)
  )
  for (i in seq_len(nrow(contrast_table))) {
    contrast_matrix[i, contrast_table$numerator[i]] <- 1
    contrast_matrix[i, contrast_table$denominator[i]] <- -1
  }
  contrast_matrix
}

make_weighted_contrast_matrix <- function(groups, label, weights) {
  missing_groups <- setdiff(names(weights), groups)
  if (length(missing_groups) > 0) {
    stop("Contrast group(s) not present in processed data: ", paste(missing_groups, collapse = ", "), call. = FALSE)
  }

  contrast_matrix <- matrix(
    0,
    nrow = 1,
    ncol = length(groups),
    dimnames = list(label, groups)
  )
  contrast_matrix[label, names(weights)] <- weights
  contrast_matrix
}

comparison_key <- function(label) {
  label |>
    stringr::str_replace_all("[^A-Za-z0-9]+", "-") |>
    stringr::str_remove("^-") |>
    stringr::str_remove("-$") |>
    stringr::str_to_lower()
}

comparison_file_key <- function(label) {
  label |>
    stringr::str_replace_all("[^A-Za-z0-9_-]+", "_") |>
    stringr::str_to_lower()
}

condition_key <- function(condition) {
  stringr::str_replace_all(condition, "[^A-Za-z0-9-]+", "_")
}

read_contrast_parameters <- function(contrast_parameters) {
  if (is.null(contrast_parameters) || length(contrast_parameters) == 0) {
    return(tibble(label = character(), numerator = character(), denominator = character()))
  }

  needed <- c("label", "numerator", "denominator")
  if (all(needed %in% names(contrast_parameters))) {
    return(tibble(
      label = as.character(unlist(contrast_parameters$label, use.names = FALSE)),
      numerator = as.character(unlist(contrast_parameters$numerator, use.names = FALSE)),
      denominator = as.character(unlist(contrast_parameters$denominator, use.names = FALSE))
    ))
  }

  purrr::map_dfr(contrast_parameters, function(comparison) {
    missing_values <- needed[!needed %in% names(comparison)]
    if (length(missing_values) > 0) {
      stop(
        "Each configured contrast needs label, numerator, and denominator. Missing: ",
        paste(missing_values, collapse = ", "),
        call. = FALSE
      )
    }
    tibble(
      label = as.character(comparison$label),
      numerator = as.character(comparison$numerator),
      denominator = as.character(comparison$denominator)
    )
  })
}

make_automatic_contrasts <- function(conditions, strategy = "all_pairwise") {
  conditions <- unique(as.character(conditions))
  if (length(conditions) < 2) {
    stop("At least two annotated conditions are required for comparison.", call. = FALSE)
  }
  if (!strategy %in% c("all_pairwise", "first_vs_rest")) {
    stop(
      "contrast_strategy must be 'all_pairwise' or 'first_vs_rest'.",
      call. = FALSE
    )
  }

  condition_pairs <- if (strategy == "first_vs_rest") {
    lapply(conditions[-1], function(condition) c(conditions[[1]], condition))
  } else {
    utils::combn(conditions, 2, simplify = FALSE)
  }

  purrr::map_dfr(condition_pairs, function(pair) {
    denominator <- pair[[1]]
    numerator <- pair[[2]]
    tibble(
      label = paste0(condition_key(numerator), "_vs_", condition_key(denominator)),
      numerator = numerator,
      denominator = denominator
    )
  })
}

make_weighted_contrasts <- function(groups, weighted_parameters) {
  if (is.null(weighted_parameters) || length(weighted_parameters) == 0) {
    return(NULL)
  }

  if (all(c("label", "weights") %in% names(weighted_parameters))) {
    labels <- as.character(unlist(weighted_parameters$label, use.names = FALSE))
    if (length(labels) == 1) {
      weighted_parameters <- list(weighted_parameters)
    } else {
      weight_sets <- weighted_parameters$weights
      if (length(weight_sets) != length(labels)) {
        stop("Each weighted contrast label needs one set of weights.", call. = FALSE)
      }
      weighted_parameters <- purrr::map2(
        labels,
        weight_sets,
        function(label, weights) list(label = label, weights = weights)
      )
    }
  }

  matrices <- purrr::map(weighted_parameters, function(comparison) {
    if (is.null(comparison$label) || is.null(comparison$weights)) {
      stop("Each weighted contrast needs label and weights.", call. = FALSE)
    }
    weights <- unlist(comparison$weights, use.names = TRUE)
    storage.mode(weights) <- "numeric"
    if (is.null(names(weights)) || any(!nzchar(names(weights)))) {
      stop("Every weighted-contrast value must be named with a condition.", call. = FALSE)
    }
    if (abs(sum(weights)) > sqrt(.Machine$double.eps)) {
      stop("Weighted contrast coefficients must sum to zero.", call. = FALSE)
    }
    make_weighted_contrast_matrix(groups, as.character(comparison$label), weights)
  })
  do.call(rbind, matrices)
}

run_group_comparison <- function(processed_data, contrast_table) {
  groups <- processed_groups(processed_data)
  contrast_matrix <- make_contrast_matrix(groups, contrast_table)

  run_group_comparison_matrix(processed_data, contrast_matrix)
}

run_group_comparison_matrix <- function(processed_data, contrast_matrix) {
  comparison <- MSstats::groupComparison(
    contrast.matrix = contrast_matrix,
    data = processed_data,
    use_log_file = FALSE,
    verbose = FALSE,
    numberOfCores = params$number_of_cores
  )

  list(
    contrast_matrix = contrast_matrix,
    comparison = comparison,
    results = as_plain_tibble(comparison[["ComparisonResult"]])
  )
}

choose_break_interval <- function(values, side, quantile_prob, min_ratio) {
  values <- values[is.finite(values)]
  if (length(values) == 0) return(NULL)

  abs_values <- abs(values)
  core_limit <- max(
    stats::quantile(abs_values, quantile_prob, na.rm = TRUE, names = FALSE),
    params$volcano_fc_cutoff * 1.5,
    2
  )
  max_abs <- max(abs_values, na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs < core_limit * min_ratio) return(NULL)

  if (side == "positive") {
    extreme_edge <- min(values[values > core_limit], na.rm = TRUE)
    if (!is.finite(extreme_edge) || extreme_edge <= core_limit) return(NULL)
    c(core_limit, extreme_edge * 0.9)
  } else {
    extreme_edge <- max(values[values < -core_limit], na.rm = TRUE)
    if (!is.finite(extreme_edge) || extreme_edge >= -core_limit) return(NULL)
    c(extreme_edge * 0.9, -core_limit)
  }
}

choose_y_break_interval <- function(neg_log10_p, quantile_prob, min_ratio) {
  neg_log10_p <- neg_log10_p[is.finite(neg_log10_p)]
  if (length(neg_log10_p) == 0) return(NULL)

  core_limit <- max(
    stats::quantile(neg_log10_p, quantile_prob, na.rm = TRUE, names = FALSE),
    -log10(params$volcano_p_cutoff) * 1.5,
    5
  )
  max_y <- max(neg_log10_p, na.rm = TRUE)
  if (!is.finite(max_y) || max_y < core_limit * min_ratio) return(NULL)

  extreme_edge <- min(neg_log10_p[neg_log10_p > core_limit], na.rm = TRUE)
  if (!is.finite(extreme_edge) || extreme_edge <= core_limit) return(NULL)

  c(core_limit, extreme_edge * 0.9)
}

apply_split_axes <- function(plot, plot_data, fc_col, p_col) {
  split_scales <- params$volcano_split_axis_scales
  split_quantile <- params$volcano_split_axis_quantile
  split_min_ratio <- params$volcano_split_axis_min_ratio

  x_values <- plot_data[[fc_col]]
  x_neg_break <- choose_break_interval(x_values, "negative", split_quantile, split_min_ratio)
  x_pos_break <- choose_break_interval(x_values, "positive", split_quantile, split_min_ratio)
  y_break <- choose_y_break_interval(-log10(plot_data[[p_col]]), split_quantile, split_min_ratio)

  if (!is.null(x_neg_break) && diff(x_neg_break) > 0) {
    plot <- plot + ggbreak::scale_x_break(x_neg_break, scales = split_scales)
  }
  if (!is.null(x_pos_break) && diff(x_pos_break) > 0) {
    plot <- plot + ggbreak::scale_x_break(x_pos_break, scales = split_scales)
  }
  if (!is.null(y_break) && diff(y_break) > 0) {
    plot <- plot + ggbreak::scale_y_break(y_break, scales = split_scales)
  }

  plot
}

plot_volcano <- function(results, comparison_label, p_col = "adj.pvalue", fc_col = "log2FC") {
  plot_data <- results |>
    filter(.data$Label == comparison_label) |>
    mutate(
      .volcano_p = pmax(.data[[p_col]], .Machine$double.xmin)
    ) |>
    filter(is.finite(.data[[fc_col]]), is.finite(.data$.volcano_p))

  plot_data <- plot_data |>
    mutate(
      GeneSymbol = map_uniprot_to_gene_symbol(.data$Protein),
      .volcano_label = if_else(
        !is.na(.data$GeneSymbol) & nzchar(.data$GeneSymbol),
        .data$GeneSymbol,
        ""
      )
    )

  top_labels <- plot_data |>
    filter(
      nzchar(.data$.volcano_label),
      .data$.volcano_p < params$volcano_p_cutoff
    ) |>
    arrange(.data$.volcano_p, desc(abs(.data[[fc_col]]))) |>
    slice_head(n = params$volcano_top_labels) |>
    pull(.data$.volcano_label)

  if (length(top_labels) == 0) {
    top_labels <- plot_data |>
      filter(nzchar(.data$.volcano_label)) |>
      arrange(.data$.volcano_p, desc(abs(.data[[fc_col]]))) |>
      slice_head(n = params$volcano_top_labels) |>
      pull(.data$.volcano_label)
  }

  volcano_plot <- EnhancedVolcano::EnhancedVolcano(
    plot_data,
    lab = plot_data$.volcano_label,
    x = fc_col,
    y = ".volcano_p",
    selectLab = top_labels,
    pCutoff = params$volcano_p_cutoff,
    FCcutoff = params$volcano_fc_cutoff,
    title = gsub(
      "_", " ",
      sub("_vs_", "/", comparison_label, fixed = TRUE),
      fixed = TRUE
    ),
    subtitle = "MSstats protein-level comparison",
    caption = paste0(
      "Adjusted p < ", params$volcano_p_cutoff,
      "; |log2FC| > ", params$volcano_fc_cutoff
    ),
    xlab = "Log2 fold change",
    ylab = "-Log10 adjusted p-value",
    pointSize = 2.2,
    labSize = 4.0,
    col = c("grey78", "grey45", "#4C72B0", "#C44E52"),
    colAlpha = 0.82,
    legendLabels = c(
      "Not significant",
      "Meets absolute log2 fold-change threshold only",
      "Meets adjusted p-value threshold only",
      "Meets both adjusted p-value and fold-change thresholds"
    ),
    legendPosition = "bottom",
    legendLabSize = 9,
    legendIconSize = 3.5,
    drawConnectors = TRUE,
    widthConnectors = 0.35,
    colConnectors = "grey35"
  ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
      plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 8)),
      plot.caption = element_text(size = 9, color = "grey30", margin = margin(t = 8)),
      axis.title = element_text(face = "bold", size = 12),
      axis.text = element_text(size = 10, color = "black"),
      legend.title = element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.text = element_text(size = 9),
      legend.margin = margin(t = 8),
      plot.margin = margin(12, 18, 12, 12)
    ) +
    guides(
      colour = guide_legend(nrow = 2, byrow = TRUE),
      shape = guide_legend(nrow = 2, byrow = TRUE)
    )

  static_plot <- apply_split_axes(volcano_plot, plot_data, fc_col, ".volcano_p")
  static_file <- file.path(
    figure_dir,
    paste0(params$request, "-", comparison_key(comparison_label), "-volcano.pdf")
  )
  ggplot2::ggsave(
    static_file,
    plot = static_plot,
    width = 12,
    height = 10,
    units = "in",
    bg = "white"
  )

  static_plot
}

create_dt <- function(data, download_basename, page_length = 25, filter = "top") {
  DT::datatable(
    data,
    rownames = FALSE,
    filter = filter,
    extensions = c("Buttons", "Scroller", "SearchBuilder"),
    width = "100%",
    options = list(
      dom = "QBlrtip",
      scrollX = TRUE,
      scrollY = 650,
      scrollCollapse = TRUE,
      scroller = TRUE,
      deferRender = TRUE,
      pageLength = page_length,
      lengthMenu = list(
        c(10, 25, 50, 100, -1),
        c("10", "25", "50", "100", "All")
      ),
      buttons = list(
        list(extend = "copy"),
        list(extend = "csv", filename = download_basename),
        list(extend = "excel", filename = download_basename)
      ),
      columnDefs = list(list(width = "42%", targets = 5))
    )
  )
}

format_p_values <- function(data) {
  p_cols <- intersect(
    c("adj.pvalue", "pvalue", "p_val_adj", "p_val"),
    names(data)
  )
  data |>
    dplyr::mutate(dplyr::across(
      dplyr::all_of(p_cols),
      ~ ifelse(is.na(.x), NA_character_, formatC(.x, format = "e", digits = 2))
    ))
}

prepare_de_table <- function(results, comparison_label) {
  result_data <- add_power_metrics(
    results,
    alpha = params$power_alpha,
    target_power = params$power_target,
    effect_log2fc = params$power_effect_log2fc
  ) |>
    as.data.frame()

  for (column in c(
    "issue", "MissingPercentage", "ImputationPercentage",
    "target_log2fc_power", "detectable_log2fc_at_target_power",
    "observed_effect_power"
  )) {
    if (!column %in% names(result_data)) result_data[[column]] <- NA
  }

  result_data |>
    as_tibble() |>
    filter(
      .data$Label == comparison_label,
      is.finite(.data$log2FC),
      is.finite(.data$pvalue),
      is.finite(.data$adj.pvalue),
      .data$adj.pvalue < params$volcano_p_cutoff,
      abs(.data$log2FC) > params$volcano_fc_cutoff
    ) |>
    mutate(
      Accession = .data$Protein,
      Gene = map_uniprot_to_gene_symbol(.data$Protein),
      Note = purrr::pmap_chr(
        list(
          .data$issue, .data$MissingPercentage, .data$ImputationPercentage,
          .data$Gene, .data$target_log2fc_power,
          .data$detectable_log2fc_at_target_power, .data$observed_effect_power
        ),
        function(issue, missing, imputation, gene, target_power, detectable_effect,
                 observed_power) {
          flags <- character()
          issue_text <- trimws(as.character(issue))
          if (!is.na(issue_text) && nzchar(issue_text) &&
              !tolower(issue_text) %in% c("na", "none", "false", "0")) {
            flags <- c(flags, paste0("MSstats flag: ", issue_text))
          }
          missing_value <- suppressWarnings(as.numeric(missing))
          if (is.finite(missing_value) && missing_value > 0) {
            flags <- c(flags, paste0("Missing measurements: ", round(missing_value, 1), "%"))
          }
          imputation_value <- suppressWarnings(as.numeric(imputation))
          if (is.finite(imputation_value) && imputation_value > 0) {
            flags <- c(flags, paste0("Imputed measurements: ", round(imputation_value, 1), "%"))
          }
          if (is.na(gene) || !nzchar(gene)) {
            flags <- c(flags, "Gene symbol unavailable")
          }
          target_power_value <- suppressWarnings(as.numeric(target_power))
          detectable_effect_value <- suppressWarnings(as.numeric(detectable_effect))
          if (is.finite(target_power_value) && target_power_value < params$power_target) {
            power_note <- paste0(
              "Power caution: ", scales::percent(target_power_value, accuracy = 1),
              " power for target |log2FC| = ", abs(params$power_effect_log2fc),
              " (below ", scales::percent(params$power_target, accuracy = 1), ")"
            )
            if (is.finite(detectable_effect_value)) {
              power_note <- paste0(
                power_note, "; estimated minimum |log2FC| = ",
                round(detectable_effect_value, 2)
              )
            }
            flags <- c(flags, power_note)
          }
          observed_power_value <- suppressWarnings(as.numeric(observed_power))
          if (is.finite(observed_power_value) && observed_power_value < params$power_target) {
            flags <- c(
              flags,
              paste0(
                "Observed-effect detectability is ",
                scales::percent(observed_power_value, accuracy = 1),
                "; interpret cautiously (not independent evidence)"
              )
            )
          }
          if (length(flags) == 0) "No additional flags." else paste(flags, collapse = "; ")
        }
      )
    ) |>
    arrange(.data$adj.pvalue, desc(abs(.data$log2FC))) |>
    select(Accession, Gene, log2FC, pvalue, adj.pvalue, Note) |>
    format_p_values()
}

as_plain_tibble <- function(x) {
  tibble::as_tibble(as.data.frame(x))
}

condition_column <- function(data) {
  candidates <- c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL")
  found <- intersect(candidates, names(data))
  if (length(found) == 0) {
    stop(
      "Could not find a condition/group column. Available columns: ",
      paste(names(data), collapse = ", "),
      call. = FALSE
    )
  }
  found[1]
}

processed_groups <- function(processed_data) {
  protein_level <- as_plain_tibble(processed_data[["ProteinLevelData"]])
  group_col <- condition_column(protein_level)
  sort(unique(as.character(protein_level[[group_col]])))
}

protein_level_group_summary <- function(processed_data) {
  protein_level <- as_plain_tibble(processed_data[["ProteinLevelData"]])
  group_col <- condition_column(protein_level)

  protein_level |>
    count(Condition = .data[[group_col]], name = "protein_summaries") |>
    arrange(.data$Condition)
}

write_results <- function(results, file_name) {
  output_file <- file.path(output_dir, file_name)
  readr::write_csv(results, output_file)
  output_file
}

quietly_run <- function(expr) {
  result <- NULL
  invisible(utils::capture.output(
    invisible(utils::capture.output(result <- force(expr), type = "message")),
    type = "output"
  ))
  result
}
```

## MSstats Processing QC

These diagnostics evaluate the converted and processed MSstats data before group
comparisons are interpreted.

```{r convert-psm-data}
msstats_input <- quietly_run(convert_pd_to_msstats(psm_filtered, annotation))

as_plain_tibble(msstats_input) |>
  count(Condition, Run, name = "features") |>
  arrange(Condition, Run) |>
  head(20) |>
  knitr::kable(
    caption = "Converted MSstats feature counts by condition and run (first 20 rows)."
  )
```

```{r process-msstats-data}
processed <- quietly_run(process_msstats(msstats_input))

protein_level_group_summary(processed) |>
  knitr::kable(
    caption = "MSstats protein-level measurement summary by experimental condition."
  )
```

```{r normalization-diagnostic, fig.height=8, fig.cap="Sample intensity distributions before and after MSstats normalization."}
normalization_distribution <- bind_rows(
  extract_distribution_data(msstats_input, "Before normalization"),
  extract_normalized_distribution_data(msstats_input, params$normalization)
)

if (nrow(normalization_distribution) > 0) {
  normalization_distribution |>
    left_join(
      annotation |> select(.data$Run, .data$FigureLabel),
      by = "Run"
    ) |>
    mutate(FigureLabel = coalesce(.data$FigureLabel, .data$Run)) |>
    ggplot(aes(x = reorder(.data$FigureLabel, .data$log2Intensity, median),
               y = .data$log2Intensity, fill = .data$Stage)) +
    geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.3) +
    coord_flip() +
    facet_wrap(~ Stage, scales = "free_x") +
    labs(x = "Sample", y = "log2 intensity",
         title = "Intensity distributions before and after MSstats normalization") +
    theme_bw()
} else {
  message("Normalization diagnostic skipped: intensity columns were not found.")
}
```

```{r processed-sample-annotation, include=FALSE}
protein_level_tbl <- as_plain_tibble(processed[["ProteinLevelData"]])
protein_run_col <- first_existing_col(
  protein_level_tbl,
  c("Run", "RUN", "originalRUN", "SUBJECT", "BioReplicate")
)
protein_condition_col <- first_existing_col(
  protein_level_tbl,
  c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL")
)

processed_sample_annotation <- if (!is.null(protein_run_col)) {
  processed_subjects <- protein_level_tbl |>
    transmute(
      Run = as.character(.data[[protein_run_col]]),
      Condition = if (!is.null(protein_condition_col)) {
        as.character(.data[[protein_condition_col]])
      } else {
        NA_character_
      }
    ) |>
    distinct() |>
    mutate(.run_order = suppressWarnings(as.numeric(.data$Run))) |>
    arrange(.data$Condition, is.na(.data$.run_order), .data$.run_order, .data$Run) |>
    group_by(.data$Condition) |>
    mutate(.condition_index = row_number()) |>
    ungroup()

  annotation_subjects <- annotation |>
    mutate(.bioreplicate_order = suppressWarnings(as.numeric(.data$BioReplicate))) |>
    arrange(.data$Condition, is.na(.data$.bioreplicate_order),
            .data$.bioreplicate_order, .data$BioReplicate) |>
    group_by(.data$Condition) |>
    mutate(.condition_index = row_number()) |>
    ungroup() |>
    select(.data$Condition, .data$.condition_index, .data$SampleName,
           .data$BioReplicate, .data$FigureLabel)

  processed_subjects |>
    left_join(annotation_subjects, by = c("Condition", ".condition_index")) |>
    select(Run, SampleName, BioReplicate, FigureLabel, Condition) |>
    distinct()
} else {
  annotation
}

```

```{r protein-level-correlation, fig.height=8, fig.cap="Pearson correlations among MSstats protein-level abundance profiles; samples are labeled by condition and biological replicate."}
protein_level_matrix <- make_protein_level_matrix(processed)

plot_correlation_heatmap(
  protein_level_matrix,
  processed_sample_annotation,
  title = "MSstats protein-level correlation between samples"
)
```

```{r protein-level-pca, fig.height=7, fig.cap="Principal component analysis of MSstats protein-level abundance profiles, colored by condition and labeled by biological replicate (BR#)."}
plot_pca_from_matrix(
  protein_level_matrix,
  processed_sample_annotation,
  title = "MSstats protein-level PCA"
)
```

```{r protein-level-cv-summary}
protein_col <- first_existing_col(protein_level_tbl, c("Protein", "PROTEIN", "ProteinName"))
condition_col <- first_existing_col(protein_level_tbl, c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL"))
value_col <- first_existing_col(protein_level_tbl, c("LogIntensities", "log2Intensity", "ABUNDANCE", "Intensity"))

if (!is.null(protein_col) && !is.null(condition_col) && !is.null(value_col)) {
  protein_cv <- protein_level_tbl |>
    transmute(
      Protein = as.character(.data[[protein_col]]),
      Condition = as.character(.data[[condition_col]]),
      linearIntensity = 2^coerce_log2_intensity(.data[[value_col]], value_col)
    ) |>
    filter(is.finite(.data$linearIntensity), .data$linearIntensity > 0) |>
    group_by(.data$Protein, .data$Condition) |>
    summarise(
      n_samples = n(),
      mean_intensity = mean(.data$linearIntensity, na.rm = TRUE),
      sd_intensity = stats::sd(.data$linearIntensity, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(cv_percent = 100 * .data$sd_intensity / .data$mean_intensity) |>
    filter(.data$n_samples >= 2, is.finite(.data$cv_percent))

  protein_cv |>
    group_by(.data$Condition) |>
    summarise(
      proteins = n(),
      median_cv_percent = median(.data$cv_percent, na.rm = TRUE),
      p75_cv_percent = stats::quantile(.data$cv_percent, 0.75, na.rm = TRUE),
      p90_cv_percent = stats::quantile(.data$cv_percent, 0.90, na.rm = TRUE),
      .groups = "drop"
    ) |>
    knitr::kable(
      digits = 2,
      caption = "Within-condition protein abundance coefficient-of-variation summary."
    )
} else {
  protein_cv <- tibble()
  message("Protein-level CV skipped: required protein, condition, or intensity columns were not found.")
}
```

```{r protein-level-cv-plot, fig.height=6, fig.cap="Distributions of within-condition protein-level coefficients of variation."}
if (nrow(protein_cv) > 0) {
  ggplot(protein_cv, aes(x = .data$Condition, y = .data$cv_percent,
                         fill = .data$Condition)) +
    geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
    geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
    labs(x = NULL, y = "CV (%)",
         title = "Protein-level coefficient of variation within condition") +
    theme_bw()
}
```

```{r top-variable-proteins}
if (nrow(protein_cv) > 0) {
  protein_cv |>
    arrange(desc(.data$cv_percent)) |>
    mutate(
      `Gene names` = map_uniprot_to_gene_symbol(.data$Protein),
      .after = Protein
    ) |>
    select(Protein, `Gene names`, Condition, n_samples, cv_percent) |>
    slice_head(n = params$qc_top_cv_proteins) |>
    knitr::kable(
      digits = 2,
      caption = "Proteins with the highest within-condition coefficients of variation."
    )
}
```

## MSstats Comparisons

The comparison names and directions below come from an optional contrast file
or from the parameter file. When neither is supplied, the report creates them
from the order of the conditions in the matched annotation rows.

```{r compare-conditions}
groups <- processed_groups(processed)
condition_order <- annotation |>
  distinct(.data$Condition) |>
  pull(.data$Condition) |>
  intersect(groups)

has_contrast_file <- !is.null(params$contrast_file) &&
  length(params$contrast_file) == 1L &&
  !is.na(params$contrast_file) &&
  nzchar(params$contrast_file)

if (has_contrast_file) {
  has_inline_contrasts <-
    (!is.null(params$contrasts) && length(params$contrasts) > 0L) ||
    (!is.null(params$weighted_contrasts) && length(params$weighted_contrasts) > 0L)
  if (has_inline_contrasts) {
    stop(
      "Use either contrast_file or inline contrasts, not both.",
      call. = FALSE
    )
  }
  resolved_contrast_file <- normalizePath(
    resolve_path(params$contrast_file), winslash = "/", mustWork = TRUE
  )
  contrast_definitions <- read_contrast_file(
    resolved_contrast_file,
    allowed_conditions = groups
  )
  all_contrast_matrix <- contrast_table_to_matrix(
    contrast_definitions,
    conditions = groups
  )
} else {
  comparison_contrasts <- read_contrast_parameters(params$contrasts)
  if (nrow(comparison_contrasts) == 0) {
    comparison_contrasts <- make_automatic_contrasts(
      condition_order,
      params$contrast_strategy
    )
  }

  pairwise_contrast_matrix <- make_contrast_matrix(groups, comparison_contrasts)
  weighted_contrast_matrix <- make_weighted_contrasts(
    groups,
    params$weighted_contrasts
  )
  all_contrast_matrix <- rbind(pairwise_contrast_matrix, weighted_contrast_matrix)
  contrast_definitions <- contrast_matrix_to_table(all_contrast_matrix)
}

contrast_guide <- describe_contrasts(contrast_definitions)
contrast_guide_file <- write_results(
  contrast_guide,
  "msstats_contrast_guide.csv"
)

if (anyDuplicated(rownames(all_contrast_matrix))) {
  stop("Every comparison label must be unique.", call. = FALSE)
}
if (anyDuplicated(purrr::map_chr(rownames(all_contrast_matrix), comparison_key))) {
  stop("Comparison labels must also be unique after filename cleanup.", call. = FALSE)
}

all_comparison_results <- quietly_run(
  run_group_comparison_matrix(processed, all_contrast_matrix)
)
all_comparison_output <- all_comparison_results$results

saveRDS(
  list(
    msstats_input = msstats_input,
    processed_data = processed[["ProcessedData"]],
    protein_level_data = processed[["ProteinLevelData"]],
    annotation = annotation,
    comparison_results = all_comparison_output,
    contrast_matrix = all_contrast_matrix,
    contrast_definitions = contrast_definitions,
    contrast_guide = contrast_guide,
    analysis_settings = list(
      normalization = params$normalization,
      summary_method = params$summary_method,
      model_based_imputation = params$mb_impute,
      number_of_cores = params$number_of_cores,
      differential_fdr_cutoff = params$volcano_p_cutoff,
      differential_log2fc_cutoff = params$volcano_fc_cutoff,
      organism_package = params$organism_package,
      contrast_file = if (has_contrast_file) params$contrast_file else NULL
    )
  ),
  file.path(output_dir, "msstats_workbook_export_objects.rds")
)

comparison_labels <- rownames(all_contrast_matrix)
all_comparisons_file <- write_results(
  all_comparison_output,
  "msstats_all_group_comparisons.csv"
)
comparison_files <- purrr::set_names(comparison_labels) |>
  purrr::map_chr(function(comparison_label) {
    write_results(
      filter(all_comparison_output, .data$Label == comparison_label),
      paste0("msstats_", comparison_file_key(comparison_label), "_group_comparison.csv")
    )
  })

knitr::kable(
  all_contrast_matrix,
  caption = "Contrast matrix used for the MSstats comparisons."
)

knitr::kable(
  contrast_guide |>
    select(.data$ContrastType, .data$Label, .data$Formula, .data$WhatItMeasures),
  caption = "Plain-language guide to the configured contrasts."
)
```

Volcano plots are publication-quality static `EnhancedVolcano` figures. Labels
use gene symbols, connectors reduce label ambiguity, and the complete
significance legend is written below each plot. Automatic `ggbreak` split axes
keep extreme tails from compressing the main protein cloud. Each figure is also
exported as a vector PDF in the `figures` directory.

```{r comparison-panels, results='asis', fig.width=12, fig.height=10, dpi=300, out.width="100%"}
invisible(DT::datatable(NULL))
for (comparison_label in comparison_labels) {
  display_label <- format_comparison_label(comparison_label)
  figure_anchor <- paste0("fig-volcano-", comparison_key(comparison_label))
  cat("### ", display_label, "\n\n", sep = "")
  cat("<span id=\"", figure_anchor, "\"></span>\n\n", sep = "")
  print(plot_volcano(all_comparison_output, comparison_label))
  cat("\n\n#### Differentially expressed proteins\n\n")
  comparison_table <- prepare_de_table(
    all_comparison_output,
    comparison_label
  ) |>
    create_dt(
      paste0(params$request, "-", comparison_key(comparison_label), "-DE-proteins")
    )
  print(htmltools::tagList(comparison_table))
  cat("\n\n")
}
```

## Gene Set Enrichment Analysis

GO Biological Process GSEA follows the same workflow for every comparison.
UniProt accessions are mapped to Entrez IDs with `r params$organism_package`,
then ranked as `sign(log2FC) * -log10(p-value)`. Duplicate Entrez mappings retain
the protein with the largest absolute ranking score. Enrichment uses
Benjamini-Hochberg correction and includes gene sets containing
`r params$gsea_min_gene_set_size` to `r params$gsea_max_gene_set_size` genes.
Positive normalized enrichment scores indicate enrichment toward proteins with
positive MSstats fold changes; negative scores indicate the reverse.

```{r run-go-bp-gsea}
gsea_results <- if (isTRUE(params$run_gsea)) {
  purrr::set_names(comparison_labels) |>
    purrr::map(function(comparison_label) {
      run_go_bp_gsea(
        all_comparison_output,
        comparison_label = comparison_label,
        min_gene_set_size = params$gsea_min_gene_set_size,
        max_gene_set_size = params$gsea_max_gene_set_size,
        pvalue_cutoff = params$gsea_pvalue_cutoff,
        dot_show_categories = params$gsea_dot_show_categories,
        tree_show_categories = params$gsea_tree_show_categories,
        tree_label_wrap = params$gsea_tree_label_wrap,
        org_db = organism_db
      )
    })
} else {
  purrr::set_names(comparison_labels) |>
    purrr::map(function(comparison_label) {
      list(
        comparison = comparison_label,
        input_proteins = 0L,
        input_accessions = 0L,
        mapped_accessions = 0L,
        mapped_entrez = 0L,
        tested_terms = 0L,
        significant_terms = 0L,
        dotplot = NULL,
        treeplot = NULL,
        dotplot_status = "Skipped",
        treeplot_status = "Skipped",
        table = empty_gsea_table(),
        status = "Skipped by report parameter."
      )
    })
}

gsea_run_summary <- summarize_gsea_runs(gsea_results)
gsea_run_summary_file <- write_results(gsea_run_summary, "gsea_go_bp_run_summary.csv")

gsea_combined_table <- purrr::imap_dfr(gsea_results, function(result, comparison_label) {
  result$table |>
    mutate(Comparison = comparison_label, .before = 1)
})
gsea_combined_file <- write_results(gsea_combined_table, "gsea_go_bp_all_comparisons.csv")

gsea_result_files <- purrr::imap_chr(gsea_results, function(result, comparison_label) {
  write_results(
    result$table,
    paste0("gsea_go_bp_", comparison_file_key(comparison_label), ".csv")
  )
})

knitr::kable(
  gsea_run_summary,
  caption = "GO Biological Process GSEA status and result counts by comparison."
)
```

### GSEA Dot Plots

<span id="fig-gsea-dotplots"></span>

```{r fig-gsea-dotplots, results='asis', fig.width=16, fig.height=10, fig.cap="GO Biological Process GSEA dot plots for each comparison. Point size represents gene count and color represents the adjusted p-value."}
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (!is.null(result$dotplot)) {
    print(result$dotplot)
  } else {
    cat("*", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```

### GSEA Tree Plots

<span id="fig-gsea-treeplots"></span>

Tree plots group enriched GO terms by pairwise term similarity, making redundant
or closely related biological processes easier to recognize.

```{r fig-gsea-treeplots, results='asis', fig.width=18, fig.height=14, fig.cap="GO Biological Process similarity trees for each comparison; branch colors group related enriched pathways."}
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (!is.null(result$treeplot)) {
    print(result$treeplot)
  } else {
    cat("*Tree plot unavailable. ", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```

### GSEA Results Tables

`NES` is the normalized enrichment score, `p.adjust` is the
Benjamini-Hochberg-adjusted p-value, and `core_enrichment` lists the leading-edge
genes driving the enrichment score.

```{r gsea-results-tables, results='asis'}
invisible(DT::datatable(NULL))
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (nrow(result$table) > 0) {
    table_widget <- DT::datatable(
      result$table,
      rownames = FALSE,
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left; font-weight: bold;",
        paste(
          "GO Biological Process GSEA results:",
          format_comparison_label(comparison_label)
        )
      ),
      extensions = "Buttons",
      options = list(
        dom = "lBfrtip",
        buttons = list(
          "copy",
          "csv",
          list(
            extend = "excel",
            filename = paste0(
              params$request, "-GSEA-",
              comparison_key(comparison_label)
            ),
            title = paste(
              "GO Biological Process GSEA results:",
              format_comparison_label(comparison_label)
            )
          )
        ),
        pageLength = 10,
        lengthMenu = list(c(10, 25, 50, -1), c("10", "25", "50", "All")),
        scrollX = TRUE
      )
    ) |>
      DT::formatSignif(
        intersect(c("enrichmentScore", "NES", "pvalue", "p.adjust", "qvalue"), names(result$table)),
        digits = 3
      )
    print(htmltools::tagList(table_widget))
  } else {
    cat("*", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```


## Power and Detectability

Power diagnostics are calculated from the MSstats contrast estimates, standard
errors, and degrees of freedom. The prospective-style calculation asks whether
the current design and observed variability would have `r scales::percent(params$power_target)`
power to detect a target effect of `r abs(params$power_effect_log2fc)` log2 fold
change at a nominal alpha of `r params$power_alpha`. The observed-effect
post-hoc power calculation uses the measured log2 fold change, so it should be
read as a detectability diagnostic alongside the p-values rather than as
independent evidence.

```{r power-settings}
tibble(
  metric = c(
    "Nominal alpha used for t-test power",
    "FDR detection threshold",
    "Target power",
    "Target absolute log2 fold-change",
    "Target linear fold-change"
  ),
  value = c(
    params$power_alpha,
    params$power_alpha,
    params$power_target,
    abs(params$power_effect_log2fc),
    round(2^abs(params$power_effect_log2fc), 3)
  )
) |>
  knitr::kable(caption = "Power-analysis settings used for all comparisons.")
```

```{r power-diagnostics}
power_diagnostics <- add_power_metrics(
  all_comparison_output,
  alpha = params$power_alpha,
  target_power = params$power_target,
  effect_log2fc = params$power_effect_log2fc
)

power_diagnostics_file <- write_results(power_diagnostics, "msstats_power_diagnostics.csv")

power_summary <- summarize_power_metrics(power_diagnostics)
power_summary |>
  knitr::kable(
    digits = 2,
    caption = "Power and minimum-detectable-effect summary by comparison."
  )
```

```{r power-detection-status}
power_diagnostics |>
  count(.data$Label, .data$power_detection_status, name = "proteins") |>
  group_by(.data$Label) |>
  mutate(percent = 100 * .data$proteins / sum(.data$proteins)) |>
  ungroup() |>
  arrange(.data$Label, desc(.data$proteins)) |>
  knitr::kable(
    digits = 2,
    caption = "Protein counts by power and differential-detection status."
  )
```

```{r fig-power-target-effect, fig.width=10, fig.height=6, fig.cap="Estimated power to detect the configured target log2 fold change across proteins in each comparison."}
power_diagnostics |>
  filter(is.finite(.data$target_log2fc_power)) |>
  ggplot(aes(x = .data$Label, y = .data$target_log2fc_power, fill = .data$Label)) +
  geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
  geom_hline(yintercept = params$power_target, linetype = "dashed", color = "grey25") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x = NULL,
    y = paste0("Power to detect |log2FC| = ", abs(params$power_effect_log2fc)),
    title = "Power for the target effect size by comparison"
  ) +
  theme_bw()
```

```{r fig-minimum-detectable-effect, fig.width=10, fig.height=6, fig.cap="Estimated minimum absolute log2 fold change required to attain the target power for each comparison."}
power_diagnostics |>
  filter(is.finite(.data$detectable_log2fc_at_target_power)) |>
  ggplot(aes(x = .data$Label, y = .data$detectable_log2fc_at_target_power,
             fill = .data$Label)) +
  geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
  geom_hline(yintercept = abs(params$power_effect_log2fc),
             linetype = "dashed", color = "grey25") +
  coord_flip() +
  labs(
    x = NULL,
    y = paste0("Minimum |log2FC| for ", scales::percent(params$power_target), " power"),
    title = "Estimated detectable effect size by comparison"
  ) +
  theme_bw()
```

```{r fig-post-hoc-power, fig.width=10, fig.height=6, fig.cap="Observed-effect power versus absolute log2 fold change, stratified by comparison and FDR detection status."}
power_diagnostics |>
  filter(is.finite(.data$observed_effect_power), is.finite(.data$log2FC)) |>
  ggplot(aes(x = abs(.data$log2FC), y = .data$observed_effect_power,
             color = .data$fdr_detected)) +
  geom_point(alpha = 0.45, size = 1.4) +
  geom_hline(yintercept = params$power_target, linetype = "dashed", color = "grey25") +
  geom_vline(xintercept = abs(params$power_effect_log2fc),
             linetype = "dotted", color = "grey35") +
  facet_wrap(~ Label) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = "#C44E52")) +
  labs(
    x = "Observed absolute log2 fold-change",
    y = "Observed-effect post-hoc power",
    color = paste0("FDR < ", params$power_alpha),
    title = "Observed-effect power and detected proteins"
  ) +
  theme_bw()
```

## Model Issue Summary

MSstats can annotate rows with model-fitting or comparison issues. This summary
counts those flags by comparison so they are visible before exported result
files are used downstream.

```{r model-issue-summary}
add_issue_status <- function(results, analysis_label) {
  results <- as_plain_tibble(results)
  if (!"issue" %in% names(results)) {
    results$issue <- NA_character_
  }

  results |>
    mutate(
      analysis = analysis_label,
      issue = as.character(.data$issue),
      issue = if_else(is.na(.data$issue) | !nzchar(.data$issue),
                      "No issue", .data$issue)
    )
}

comparison_issue_summary <- add_issue_status(
  all_comparison_output,
  "Configured comparisons"
) |>
  count(.data$analysis, .data$Label, .data$issue, name = "proteins") |>
  arrange(.data$analysis, .data$Label, desc(.data$proteins))

knitr::kable(
  comparison_issue_summary,
  caption = "MSstats model and comparison issue counts."
)
```

## Conclusion

```{r conclusion, results='asis'}
comparison_figure_anchors <- stats::setNames(
  paste0("#fig-volcano-", purrr::map_chr(comparison_labels, comparison_key)),
  comparison_labels
)

render_analysis_conclusion(
  all_comparison_output,
  gsea_combined_table,
  fdr_cutoff = params$volcano_p_cutoff,
  log2fc_cutoff = params$volcano_fc_cutoff,
  power_summary = power_summary,
  target_power = params$power_target,
  power_effect_log2fc = params$power_effect_log2fc,
  comparison_figure_anchors = comparison_figure_anchors
)
```

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function () {
  const conclusion = document.getElementById("conclusion");
  const methods = document.getElementById("methods");
  if (conclusion && methods && methods.parentNode) {
    methods.parentNode.insertBefore(conclusion, methods);
  }

  const tocLink = document.querySelector('#TOC a[href="#conclusion"]');
  if (tocLink) {
    const tocItem = tocLink.closest("li");
    const tocList = tocItem ? tocItem.parentNode : null;
    if (tocItem && tocList) tocList.insertBefore(tocItem, tocList.firstChild);
  }

  document.addEventListener("click", function (event) {
    const link = event.target.closest('a[href^="#fig-"]');
    if (!link) return;
    const targetId = link.getAttribute("href").slice(1);
    const target = document.getElementById(targetId);
    const tabPane = target ? target.closest(".tab-pane") : null;
    if (!target || !tabPane || tabPane.classList.contains("active")) return;

    const selector =
      '[data-bs-target="#' + tabPane.id + '"], [href="#' + tabPane.id + '"]';
    const tabControl = document.querySelector(selector);
    if (tabControl && window.bootstrap) {
      event.preventDefault();
      bootstrap.Tab.getOrCreateInstance(tabControl).show();
      window.setTimeout(function () {
        target.scrollIntoView({behavior: "smooth", block: "start"});
      }, 180);
    }
  });
});
</script>
```

## Session Info

```{r session-info}
sessionInfo()
```
