---
title: "Time-course Proteomics with maSigPro and Mfuzz"
subtitle: "Differential progression and soft clustering from MSstats protein abundances"
author:
  - Kevin Cormier
date: today
title-block-banner: "#27445C"
format:
  html:
    toc: true
    toc-location: left
    toc-active-border: red
    page-layout: full
    theme:
      light: [cosmo, assets/offline-fonts.scss]
      dark: [darkly, assets/offline-fonts.scss]
    code-fold: true
    embed-resources: true
    smooth-scroll: true
    fig-align: center
    fig-width: 10
    fig-height: 7
execute:
  cache: false
  warning: false
  message: false
params:
  request: "timecourse-analysis"
  report_label: "Time-course proteomics project"
  analysis_file: null
  output_dir: "results/timecourse-analysis"
  figures_dir: "figures/timecourse-analysis"
  time_column: "Time"
  group_column: "Group"
  reference_group: "Reference"
  comparison_group: "Treatment"
  time_unit: 24
  min_replicates_per_cell: 4
  polynomial_degree: 3
  model_fdr: 0.05
  coefficient_alpha: 0.05
  min_r_squared: 0.60
  min_clusters: 2
  max_clusters: 10
  cluster_diagnostic_repeats: 20
  core_membership: 0.70
  random_seed: 20261007
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)

required_packages <- c(
  "Biobase",
  "data.table",
  "dplyr",
  "DT",
  "ggplot2",
  "knitr",
  "maSigPro",
  "Mfuzz",
  "readr",
  "scales",
  "tibble",
  "tidyr",
  "xfun"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  macos_mfuzz_note <- if (
    identical(Sys.info()[["sysname"]], "Darwin") &&
    "Mfuzz" %in% missing_packages
  ) {
    paste0(
      " On macOS, an installed Mfuzz package may still require XQuartz ",
      "because its optional Tk interface is imported when the package loads."
    )
  } else {
    ""
  }
  stop(
    "Install required package(s): ",
    paste(missing_packages, collapse = ", "),
    ". maSigPro and Mfuzz are available from Bioconductor.",
    macos_mfuzz_note,
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(Biobase)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(maSigPro)
  library(Mfuzz)
  library(tibble)
  library(tidyr)
})

current_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
report_dir <- if (!is.null(current_input) && nzchar(current_input)) {
  dirname(normalizePath(current_input, winslash = "/", mustWork = TRUE))
} else {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}

required_value <- function(value, parameter_name) {
  if (is.null(value) || length(value) != 1L || is.na(value) || !nzchar(value)) {
    stop("Set the ", parameter_name, " parameter before rendering.", call. = FALSE)
  }
  value
}

resolve_path <- function(path) {
  path <- required_value(path, "file path")
  if (grepl("^(/|[A-Za-z]:[/\\\\])", path)) path else file.path(report_dir, path)
}

analysis_file <- normalizePath(
  resolve_path(required_value(params$analysis_file, "analysis_file")),
  winslash = "/",
  mustWork = TRUE
)
output_dir <- resolve_path(params$output_dir)
figure_dir <- resolve_path(params$figures_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

# A relative figure path lets Pandoc embed figures even when figures_dir was
# supplied as an absolute path.
report_figure_dir <- xfun::relative_path(
  normalizePath(figure_dir, winslash = "/", mustWork = TRUE),
  dir = report_dir
)
knitr::opts_chunk$set(
  fig.path = file.path(report_figure_dir, paste0(params$request, "-")),
  dev = c("png", "pdf")
)

set.seed(as.integer(params$random_seed))

write_result <- function(data, filename) {
  path <- file.path(output_dir, filename)
  readr::write_csv(as.data.frame(data), path, na = "")
  normalizePath(path, winslash = "/", mustWork = TRUE)
}

extract_significant_names <- function(significant_object) {
  profiles <- significant_object$sig.profiles
  if (is.null(profiles) || !nrow(profiles)) character() else rownames(profiles)
}

choose_cluster_count <- function(cluster_numbers, minimum_distances, nonempty_matrix) {
  fully_nonempty <- apply(nonempty_matrix, 2, min) == cluster_numbers
  eligible <- which(fully_nonempty)
  if (length(eligible) < 3L) {
    return(cluster_numbers[max(eligible, 1L)])
  }

  x <- cluster_numbers[eligible]
  y <- minimum_distances[eligible]
  x_scaled <- (x - min(x)) / (max(x) - min(x))
  y_scaled <- (y - min(y)) / (max(y) - min(y))
  straight_line <- 1 - x_scaled
  elbow_score <- straight_line - y_scaled
  x[[which.max(elbow_score)]]
}
```

```{r prepare-data, include=FALSE}
analysis_objects <- readRDS(analysis_file)
required_objects <- c("protein_level_data", "annotation", "analysis_settings")
missing_objects <- setdiff(required_objects, names(analysis_objects))
if (length(missing_objects)) {
  stop(
    "The analysis RDS is missing: ", paste(missing_objects, collapse = ", "),
    call. = FALSE
  )
}

annotation_source <- as_tibble(analysis_objects$annotation)
required_annotation_columns <- c(
  "Run", "BioReplicate", params$time_column, params$group_column
)
missing_annotation_columns <- setdiff(
  required_annotation_columns, names(annotation_source)
)
if (length(missing_annotation_columns)) {
  stop(
    "The saved annotation is missing: ",
    paste(missing_annotation_columns, collapse = ", "),
    call. = FALSE
  )
}

groups <- c(params$reference_group, params$comparison_group)
annotation <- annotation_source |>
  transmute(
    Run = as.character(.data$Run),
    Time = suppressWarnings(as.numeric(.data[[params$time_column]])),
    Group = as.character(.data[[params$group_column]]),
    BioReplicate = as.character(.data$BioReplicate)
  ) |>
  filter(.data$Group %in% groups) |>
  mutate(Group = factor(.data$Group, levels = groups)) |>
  arrange(.data$Time, .data$Group, .data$BioReplicate) |>
  group_by(.data$Time, .data$Group) |>
  mutate(Replicate = row_number()) |>
  ungroup()

if (any(!is.finite(annotation$Time))) {
  stop("The time column must contain numeric values.", call. = FALSE)
}
if (anyDuplicated(annotation$Run)) {
  stop("Each run must occur exactly once in the saved annotation.", call. = FALSE)
}
if (anyDuplicated(annotation$BioReplicate)) {
  stop(
    "This report is configured for independent samples, but BioReplicate identifiers repeat.",
    call. = FALSE
  )
}
if (!setequal(as.character(unique(annotation$Group)), groups)) {
  stop("Both configured groups must be present in the annotation.", call. = FALSE)
}

cell_sizes <- annotation |>
  count(.data$Time, .data$Group, name = "AvailableReplicates")
if (any(cell_sizes$AvailableReplicates < params$min_replicates_per_cell)) {
  stop(
    "At least one group-time cell has fewer samples than min_replicates_per_cell.",
    call. = FALSE
  )
}

protein_source <- as_tibble(analysis_objects$protein_level_data)
required_protein_columns <- c("Protein", "originalRUN", "LogIntensities")
missing_protein_columns <- setdiff(required_protein_columns, names(protein_source))
if (length(missing_protein_columns)) {
  stop(
    "The saved protein-level table is missing: ",
    paste(missing_protein_columns, collapse = ", "),
    call. = FALSE
  )
}

protein_data <- protein_source |>
  transmute(
    Protein = as.character(.data$Protein),
    Run = as.character(.data$originalRUN),
    Log2Abundance = suppressWarnings(as.numeric(.data$LogIntensities))
  ) |>
  inner_join(annotation, by = "Run")

duplicate_protein_runs <- protein_data |>
  count(.data$Protein, .data$Run, name = "Rows") |>
  filter(.data$Rows > 1L)
if (nrow(duplicate_protein_runs)) {
  stop("Protein/run rows must be unique before time-course modeling.", call. = FALSE)
}

time_values <- sort(unique(annotation$Time))
effect_label <- paste(params$comparison_group, "-", params$reference_group)
interaction_label <- paste0(params$comparison_group, "-by-time")
group_time_cells <- length(time_values) * length(groups)
coverage <- protein_data |>
  filter(is.finite(.data$Log2Abundance)) |>
  count(.data$Protein, .data$Time, .data$Group, name = "MeasuredReplicates") |>
  complete(
    Protein,
    Time = time_values,
    Group = factor(groups, levels = groups),
    fill = list(MeasuredReplicates = 0L)
  )

protein_coverage <- coverage |>
  group_by(.data$Protein) |>
  summarise(
    MinimumCellReplicates = min(.data$MeasuredReplicates),
    TotalMeasurements = sum(.data$MeasuredReplicates),
    .groups = "drop"
  )

coverage_proteins <- protein_coverage |>
  filter(.data$MinimumCellReplicates >= params$min_replicates_per_cell) |>
  pull(.data$Protein)

sample_order <- annotation$Run
expression_table <- protein_data |>
  filter(.data$Protein %in% coverage_proteins) |>
  select(.data$Protein, .data$Run, .data$Log2Abundance) |>
  pivot_wider(names_from = .data$Run, values_from = .data$Log2Abundance)

expression_matrix <- as.matrix(expression_table[, sample_order, drop = FALSE])
storage.mode(expression_matrix) <- "double"
rownames(expression_matrix) <- expression_table$Protein
protein_sd <- apply(expression_matrix, 1, stats::sd, na.rm = TRUE)
variable_proteins <- is.finite(protein_sd) & protein_sd > 0
expression_matrix <- expression_matrix[variable_proteins, , drop = FALSE]

minimum_observations <- params$min_replicates_per_cell *
  length(time_values) * length(groups)

filter_summary <- tibble(
  Step = c(
    "Proteins in saved MSstats protein-level data",
    paste0(
      "At least ", params$min_replicates_per_cell,
      " replicates in every group-time cell"
    ),
    "Non-zero variation across samples"
  ),
  Proteins = c(
    n_distinct(protein_data$Protein),
    length(coverage_proteins),
    nrow(expression_matrix)
  )
)
```

```{r run-masigpro, include=FALSE}
baseline_time <- min(annotation$Time)
scaled_time <- (annotation$Time - baseline_time) / params$time_unit

experimental_design <- data.frame(
  Time = scaled_time,
  Replicate = annotation$Replicate,
  row.names = annotation$Run,
  check.names = FALSE
)
experimental_design[[params$reference_group]] <- as.integer(
  annotation$Group == params$reference_group
)
experimental_design[[params$comparison_group]] <- as.integer(
  annotation$Group == params$comparison_group
)

masigpro_design <- maSigPro::make.design.matrix(
  experimental_design,
  degree = params$polynomial_degree,
  time.col = 1,
  repl.col = 2,
  group.cols = 3:4
)

masigpro_console <- capture.output(
  masigpro_fit <- maSigPro::p.vector(
    expression_matrix,
    masigpro_design,
    Q = params$model_fdr,
    MT.adjust = "BH",
    min.obs = minimum_observations,
    counts = FALSE
  )
)

tfit_console <- capture.output(
  masigpro_tfit <- maSigPro::T.fit(
    masigpro_fit,
    step.method = "backward",
    alfa = params$coefficient_alpha
  )
)

significant_all <- maSigPro::get.siggenes(
  masigpro_tfit,
  rsq = params$min_r_squared,
  vars = "all"
)
significant_each <- maSigPro::get.siggenes(
  masigpro_tfit,
  rsq = params$min_r_squared,
  vars = "each"
)
significant_groups <- maSigPro::get.siggenes(
  masigpro_tfit,
  rsq = params$min_r_squared,
  vars = "groups"
)

variable_significance <- bind_rows(lapply(
  names(significant_each$sig.genes),
  function(variable) {
    tibble(
      Variable = variable,
      Protein = extract_significant_names(significant_each$sig.genes[[variable]])
    )
  }
))

variable_counts <- variable_significance |>
  count(.data$Variable, name = "Proteins") |>
  complete(
    Variable = names(significant_each$sig.genes),
    fill = list(Proteins = 0L)
  )

interaction_variables <- names(significant_each$sig.genes)[
  grepl(
    paste0("x", params$comparison_group),
    names(significant_each$sig.genes),
    fixed = TRUE
  )
]
differential_time_proteins <- variable_significance |>
  filter(.data$Variable %in% interaction_variables) |>
  distinct(.data$Protein) |>
  pull(.data$Protein)

if (!length(differential_time_proteins)) {
  stop(
    "No proteins passed the maSigPro time-by-group interaction criteria; Mfuzz cannot proceed.",
    call. = FALSE
  )
}

overall_model_results <- tibble(
  Protein = rownames(masigpro_fit$p.vector),
  OverallModelPValue = as.numeric(masigpro_fit$p.vector[, 1]),
  OverallModelAdjustedPValue = as.numeric(masigpro_fit$p.adjusted)
)

selected_model_results <- as.data.frame(
  masigpro_tfit$sol,
  check.names = FALSE
) |>
  rownames_to_column("Protein")
names(selected_model_results)[2:3] <- c(
  "StepwiseModelPValue", "ModelRSquared"
)

masigpro_results <- overall_model_results |>
  left_join(selected_model_results, by = "Protein") |>
  mutate(
    PassedOverallModelFDR = .data$OverallModelAdjustedPValue <= params$model_fdr,
    PassedRSquared = !is.na(.data$ModelRSquared) &
      .data$ModelRSquared >= params$min_r_squared,
    DifferentialTimeInteraction = .data$Protein %in% differential_time_proteins
  ) |>
  arrange(
    desc(.data$DifferentialTimeInteraction),
    .data$OverallModelAdjustedPValue,
    desc(.data$ModelRSquared)
  )

group_counts <- tibble(
  GroupComponent = names(significant_groups$sig.genes),
  Proteins = vapply(
    significant_groups$sig.genes,
    function(x) length(extract_significant_names(x)),
    integer(1)
  )
)
```

```{r run-mfuzz, include=FALSE}
condition_summary <- protein_data |>
  filter(
    .data$Protein %in% differential_time_proteins,
    is.finite(.data$Log2Abundance)
  ) |>
  group_by(.data$Protein, .data$Time, .data$Group) |>
  summarise(
    MeanLog2Abundance = mean(.data$Log2Abundance),
    SDLog2Abundance = stats::sd(.data$Log2Abundance),
    Replicates = n(),
    SELog2Abundance = .data$SDLog2Abundance / sqrt(.data$Replicates),
    .groups = "drop"
  )

effect_long <- condition_summary |>
  select(.data$Protein, .data$Time, .data$Group, .data$MeanLog2Abundance) |>
  pivot_wider(names_from = .data$Group, values_from = .data$MeanLog2Abundance) |>
  mutate(
    Effect = .data[[params$comparison_group]] -
      .data[[params$reference_group]]
  )

effect_table <- effect_long |>
  select(.data$Protein, .data$Time, .data$Effect) |>
  pivot_wider(names_from = .data$Time, values_from = .data$Effect)
time_columns <- as.character(time_values)
effect_matrix <- as.matrix(effect_table[, time_columns, drop = FALSE])
storage.mode(effect_matrix) <- "double"
rownames(effect_matrix) <- effect_table$Protein

if (any(!is.finite(effect_matrix))) {
  stop(
    "The balanced coverage filter should produce complete condition means, but missing effects remain.",
    call. = FALSE
  )
}

effect_eset <- Biobase::ExpressionSet(
  assayData = effect_matrix,
  phenoData = Biobase::AnnotatedDataFrame(data.frame(
    Time = time_values,
    row.names = time_columns
  ))
)

# No values should be removed here; these calls make the Mfuzz preprocessing
# explicit and guard against unexpected missing or constant effect profiles.
mfuzz_filter_log <- capture.output(
  effect_eset <- Mfuzz::filter.NA(effect_eset, thres = 0)
)
mfuzz_sd_log <- capture.output(
  effect_eset <- Mfuzz::filter.std(effect_eset, min.std = 0, visu = FALSE)
)
standardized_eset <- Mfuzz::standardise(effect_eset)
fuzzifier <- Mfuzz::mestimate(standardized_eset)

maximum_candidate <- min(
  as.integer(params$max_clusters),
  nrow(Biobase::exprs(standardized_eset)) - 1L
)
cluster_candidates <- seq.int(
  as.integer(params$min_clusters),
  maximum_candidate
)
if (length(cluster_candidates) < 2L) {
  stop("At least two candidate cluster counts are required.", call. = FALSE)
}

set.seed(as.integer(params$random_seed))
minimum_centroid_distance <- Mfuzz::Dmin(
  standardized_eset,
  m = fuzzifier,
  crange = cluster_candidates,
  repeats = as.integer(params$cluster_diagnostic_repeats),
  visu = FALSE
)
set.seed(as.integer(params$random_seed))
nonempty_clusters <- Mfuzz::cselection(
  standardized_eset,
  m = fuzzifier,
  crange = cluster_candidates,
  repeats = as.integer(params$cluster_diagnostic_repeats),
  visu = FALSE
)

chosen_clusters <- choose_cluster_count(
  cluster_candidates,
  minimum_centroid_distance,
  nonempty_clusters
)

set.seed(as.integer(params$random_seed))
mfuzz_fit <- Mfuzz::mfuzz(
  standardized_eset,
  centers = chosen_clusters,
  m = fuzzifier,
  iter.max = 1000
)

membership_table <- as.data.frame(mfuzz_fit$membership) |>
  rownames_to_column("Protein")
names(membership_table)[-1] <- paste0(
  "MembershipCluster", seq_len(chosen_clusters)
)
membership_matrix <- as.matrix(membership_table[, -1, drop = FALSE])

cluster_assignments <- membership_table |>
  mutate(
    Cluster = as.integer(mfuzz_fit$cluster),
    MaximumMembership = apply(membership_matrix, 1, max),
    CoreMember = .data$MaximumMembership >= params$core_membership
  ) |>
  relocate(
    .data$Protein, .data$Cluster, .data$MaximumMembership,
    .data$CoreMember
  )

centroid_table <- as.data.frame(mfuzz_fit$centers) |>
  rownames_to_column("Cluster") |>
  mutate(Cluster = as.integer(.data$Cluster))
names(centroid_table)[-1] <- paste0("Time", time_values)

cluster_diagnostics <- tibble(
  CandidateClusters = cluster_candidates,
  MinimumCentroidDistance = minimum_centroid_distance,
  MedianNonemptyClusters = apply(nonempty_clusters, 2, stats::median),
  MinimumNonemptyClusters = apply(nonempty_clusters, 2, min),
  MaximumNonemptyClusters = apply(nonempty_clusters, 2, max),
  Selected = cluster_candidates == chosen_clusters
)

cluster_summary <- cluster_assignments |>
  group_by(.data$Cluster) |>
  summarise(
    Proteins = n(),
    CoreProteins = sum(.data$CoreMember),
    MedianMaximumMembership = stats::median(.data$MaximumMembership),
    .groups = "drop"
  ) |>
  left_join(
    effect_long |>
      inner_join(
        cluster_assignments |>
          select(.data$Protein, .data$Cluster),
        by = "Protein"
      ) |>
      group_by(.data$Cluster, .data$Time) |>
      summarise(MedianEffect = stats::median(.data$Effect), .groups = "drop") |>
      group_by(.data$Cluster) |>
      summarise(
        PeakPositiveTime = .data$Time[which.max(.data$MedianEffect)],
        PeakNegativeTime = .data$Time[which.min(.data$MedianEffect)],
        EarlyMedianEffect = .data$MedianEffect[which.min(.data$Time)],
        LateMedianEffect = .data$MedianEffect[which.max(.data$Time)],
        .groups = "drop"
      ),
    by = "Cluster"
  )
```

```{r save-results, include=FALSE}
filter_file <- write_result(protein_coverage, "timecourse_protein_coverage.csv")
design_file <- write_result(
  rownames_to_column(as.data.frame(masigpro_design$dis), "Run"),
  "masigpro_design_matrix.csv"
)
model_file <- write_result(masigpro_results, "masigpro_model_results.csv")
variable_file <- write_result(
  variable_significance,
  "masigpro_significant_proteins_by_variable.csv"
)
condition_file <- write_result(
  condition_summary,
  "timecourse_condition_means.csv"
)
effect_file <- write_result(effect_long, "timecourse_group_difference_effects.csv")
membership_file <- write_result(
  cluster_assignments,
  "mfuzz_cluster_membership.csv"
)
centroid_file <- write_result(centroid_table, "mfuzz_cluster_centroids.csv")
cluster_summary_file <- write_result(
  cluster_summary,
  "mfuzz_cluster_summary.csv"
)
diagnostic_file <- write_result(
  cluster_diagnostics,
  "mfuzz_cluster_number_diagnostics.csv"
)

saveRDS(
  list(
    filtering = list(
      coverage = protein_coverage,
      retained_proteins = rownames(expression_matrix),
      minimum_replicates_per_cell = params$min_replicates_per_cell
    ),
    maSigPro = list(
      experimental_design = experimental_design,
      regression_design = masigpro_design,
      p_vector = masigpro_fit,
      t_fit = masigpro_tfit,
      significant_by_variable = variable_significance,
      differential_time_proteins = differential_time_proteins
    ),
    Mfuzz = list(
      effect_expression_set = effect_eset,
      standardized_expression_set = standardized_eset,
      fuzzifier = fuzzifier,
      cluster_diagnostics = cluster_diagnostics,
      selected_clusters = chosen_clusters,
      fit = mfuzz_fit,
      assignments = cluster_assignments
    ),
    parameters = params
  ),
  file.path(output_dir, "timecourse_masigpro_mfuzz_objects.rds")
)
```

# Analysis overview

This report analyzes **`r params$report_label`** using the normalized
protein-level log2 abundances saved by the preceding MSstats report.

The design contains `r nrow(annotation)` independent samples, `r length(time_values)`
time points (`r paste(time_values, collapse = ", ")` hours), and two groups:
`r params$reference_group` and `r params$comparison_group`. Each group-time
cell contains `r paste(sort(unique(cell_sizes$AvailableReplicates)), collapse = ", ")`
biological replicates.

The two analyses answer different questions:

- **maSigPro** performs formal regression tests for time-related protein
  patterns and differences between the `r params$comparison_group` and `r
  params$reference_group` trajectories.
- **Mfuzz** performs exploratory soft clustering of the time-dependent `r
  effect_label` effect among proteins selected by maSigPro. Membership values
  describe how strongly each protein follows a cluster pattern.

# Detailed methods

## Source values and experimental unit

The input is `r basename(analysis_file)`, created by the MSstats report. The
`ProteinLevelData` values had already been log2 transformed, normalized by
`r analysis_objects$analysis_settings$normalization`, summarized with
`r analysis_objects$analysis_settings$summary_method`, and processed with
MSstats model-based imputation set to
`r analysis_objects$analysis_settings$model_based_imputation`.

No PSM-level values were reprocessed here. A row represents one protein in one
run. `BioReplicate` identifiers are unique across all runs, so samples are
independent rather than repeated measurements of the same subject.

## Coverage filtering

Time-course models can be driven by uneven missingness if a protein is abundant
in only one group or one part of the experiment. A protein was therefore
required to have a finite MSstats protein abundance in at least
`r params$min_replicates_per_cell` biological replicates in **every** one of the
`r group_time_cells` group-time cells. This per-cell coverage rule is stricter
than an overall missing-value threshold. Proteins with zero variance across all
samples were also removed.

No additional sample-level values were imputed for maSigPro. Its Gaussian
regressions used the observed finite values and required at least
`r minimum_observations` observations per protein.

## maSigPro regression

[maSigPro](https://bioconductor.org/packages/maSigPro) is a regression-based
method for detecting significant time-course profiles and differences among
experimental groups. The input was the filtered, normalized log2 protein matrix
with proteins in rows and independent samples in columns.

Time was expressed as

\[
t = \frac{\text{hour} - `r baseline_time`}{`r params$time_unit`},
\]

so the intercept corresponds to `r baseline_time` hours and one time unit is
`r params$time_unit` hours. Scaling improves numerical stability without
changing the fitted curves.

A degree-`r params$polynomial_degree` polynomial was used. With `r
length(time_values)` time points, this model can represent broad monotonic and
curved responses without fitting a separate parameter for every time point.
With `r params$reference_group` as the reference, the model is conceptually

\[
y = \beta_0 + \beta_G G +
\sum_{k=1}^{`r params$polynomial_degree`} \beta_k t^k +
\sum_{k=1}^{`r params$polynomial_degree`} \gamma_k(Gt^k) + \epsilon,
\]

where \(G=1\) for `r params$comparison_group` and 0 for `r
params$reference_group`. The \(\gamma_k\) interaction terms test whether the
`r params$comparison_group` trajectory differs from `r params$reference_group`
over time.

`p.vector()` first compared the full model with an intercept-only model using a
Gaussian F test. P-values were adjusted across proteins with Benjamini-Hochberg
and the FDR threshold was `r params$model_fdr`. `T.fit()` then used backward
variable selection with coefficient threshold `r params$coefficient_alpha`.
`get.siggenes()` retained models with \(R^2 \ge `r params$min_r_squared`\).

The set passed to Mfuzz was deliberately narrower than the overall maSigPro
set: a protein had to be selected for at least one `r interaction_label`
interaction (`r paste(interaction_variables, collapse = ", ")`). A significant
constant group offset without a time interaction was not considered
differential progression.

## Mfuzz soft clustering

[Mfuzz](https://bioconductor.org/packages/Mfuzz) uses fuzzy c-means clustering.
The package documentation recommends averaging replicates before clustering;
therefore, observed log2 abundances were averaged within each group-time cell.
For every maSigPro interaction protein, the clustering profile was

\[
\Delta_t = \overline{y}_{\text{comparison},t} -
           \overline{y}_{\text{reference},t}
\]

at the `r length(time_values)` time points. Thus, the clusters describe the
*shape of the `r effect_label` effect over time*, not merely abundance over time
in either group.

The balanced coverage filter guaranteed complete group means, so `fill.NA()`
was not used. `filter.NA(thres = 0)` and `filter.std(min.std = 0)` were retained
as safeguards. `standardise()` transformed each effect profile to mean zero and
standard deviation one because Mfuzz uses Euclidean distance. Standardization
affects cluster shape but not the log2-effect values reported in the tables.

The fuzzifier was estimated from the data with `mestimate()` and equaled
`r round(fuzzifier, 3)`. Candidate cluster counts from
`r min(cluster_candidates)` through `r max(cluster_candidates)` were evaluated
with `Dmin()` and `cselection()` using
`r params$cluster_diagnostic_repeats` repeated starts. Counts that ever produced
an empty cluster were excluded. Among the remaining counts, the selected value
was the elbow with the largest distance below the straight line joining the
first and last minimum-centroid-distance values. This selected
**`r chosen_clusters` clusters**. The final clustering used a fixed seed and up
to 1,000 iterations.

Proteins were assigned to their highest-membership cluster for summaries, but
all membership values are exported. A core member has maximum membership at
least `r params$core_membership`; this threshold is descriptive rather than a
significance test.

## Interpretation limits

- maSigPro assumes independent residuals here; it is not a repeated-measures
  model.
- Cubic polynomials summarize broad trajectories and should not be interpreted
  as mechanistic kinetic models.
- The source values include MSstats feature-level model-based imputation. The
  per-cell coverage filter reduces, but does not eliminate, dependence on that
  earlier processing.
- Backward variable selection and the interaction labels identify candidate
  trajectory differences; important proteins should be inspected in the raw
  replicate plots.
- Mfuzz clusters and memberships are exploratory. Cluster membership does not
  supply a p-value or prove a biological pathway.

# Filtering results

```{r filtering-table}
filter_summary |>
  knitr::kable(
    caption = "Protein counts after each time-course filtering step."
  )
```

```{r coverage-distribution, fig.height=5, fig.cap="Minimum number of observed biological replicates across all group-time cells for each protein."}
protein_coverage |>
  count(.data$MinimumCellReplicates, name = "Proteins") |>
  ggplot(aes(x = factor(.data$MinimumCellReplicates), y = .data$Proteins)) +
  geom_col(fill = "#4C72B0") +
  geom_vline(
    xintercept = params$min_replicates_per_cell + 0.5,
    linetype = "dashed",
    color = "#C44E52"
  ) +
  labs(
    x = "Minimum replicates observed in any group-time cell",
    y = "Proteins",
    title = "Balanced coverage filtering"
  ) +
  theme_bw()
```

# maSigPro results

The overall model identified **`r masigpro_fit$i` proteins** at BH-adjusted
FDR `r params$model_fdr`. After backward variable selection and the
\(R^2\) threshold, **`r significant_all$sig.genes$g` proteins** retained an
adequate fitted trajectory. At least one `r interaction_label` interaction was selected
for **`r length(differential_time_proteins)` proteins**; these were carried into
Mfuzz.

```{r masigpro-design}
knitr::kable(
  head(rownames_to_column(as.data.frame(masigpro_design$dis), "Run"), 12),
  digits = 3,
  caption = "First 12 rows of the maSigPro regression design matrix."
)
```

```{r masigpro-variable-counts, fig.height=6, fig.cap="Numbers of proteins selected for each maSigPro model term after the R-squared filter."}
variable_counts |>
  mutate(Variable = reorder(.data$Variable, .data$Proteins)) |>
  ggplot(aes(x = .data$Variable, y = .data$Proteins)) +
  geom_col(fill = "#4C72B0") +
  coord_flip() +
  labs(
    x = "Model term",
    y = "Selected proteins",
    title = "maSigPro variable-selection results"
  ) +
  theme_bw()
```

```{r masigpro-group-counts}
group_counts |>
  knitr::kable(
    caption = "Proteins assigned to each maSigPro group-level component."
  )
```

```{r top-interaction-trajectories, fig.height=11, fig.cap="Observed normalized protein abundances for the highest-R-squared proteins with a group-by-time interaction. Points are independent biological samples; lines connect group-time means."}
top_interaction_proteins <- masigpro_results |>
  filter(.data$DifferentialTimeInteraction) |>
  arrange(desc(.data$ModelRSquared), .data$OverallModelAdjustedPValue) |>
  slice_head(n = 12) |>
  pull(.data$Protein)

protein_data |>
  filter(.data$Protein %in% top_interaction_proteins) |>
  mutate(
    Protein = factor(.data$Protein, levels = top_interaction_proteins),
    Group = factor(.data$Group, levels = groups)
  ) |>
  ggplot(aes(x = .data$Time, y = .data$Log2Abundance, color = .data$Group)) +
  geom_point(alpha = 0.55, size = 1.2, position = position_jitter(width = 1.2)) +
  stat_summary(aes(group = .data$Group), fun = mean, geom = "line", linewidth = 0.8) +
  stat_summary(aes(group = .data$Group), fun = mean, geom = "point", size = 2) +
  facet_wrap(~ Protein, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c("#4C72B0", "#C44E52")) +
  labs(
    x = "Time (hours)",
    y = "MSstats normalized log2 abundance",
    color = NULL,
    title = "Top differential time-course proteins"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r masigpro-results-table}
DT::datatable(
  masigpro_results |>
    filter(.data$DifferentialTimeInteraction) |>
    select(
      .data$Protein,
      .data$OverallModelAdjustedPValue,
      .data$ModelRSquared,
      .data$DifferentialTimeInteraction
    ),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = "Proteins with at least one selected comparison-group-by-time interaction."
)
```

# Mfuzz results

## Cluster-number diagnostics

```{r mfuzz-dmin, fig.height=5, fig.cap="Mfuzz minimum centroid distance across candidate cluster counts. The selected elbow is marked in red."}
cluster_diagnostics |>
  ggplot(aes(
    x = .data$CandidateClusters,
    y = .data$MinimumCentroidDistance
  )) +
  geom_line(color = "#4C72B0", linewidth = 0.8) +
  geom_point(color = "#4C72B0", size = 2) +
  geom_vline(
    xintercept = chosen_clusters,
    color = "#C44E52",
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = cluster_candidates) +
  labs(
    x = "Candidate clusters",
    y = "Minimum centroid distance",
    title = paste("Dmin elbow selected", chosen_clusters, "clusters")
  ) +
  theme_bw()
```

```{r mfuzz-nonempty, fig.height=5, fig.cap="Mfuzz cselection diagnostic. All selected candidate clusters should remain non-empty over repeated starts."}
cluster_diagnostics |>
  ggplot(aes(
    x = .data$CandidateClusters,
    y = .data$MedianNonemptyClusters
  )) +
  geom_ribbon(aes(
    ymin = .data$MinimumNonemptyClusters,
    ymax = .data$MaximumNonemptyClusters
  ), fill = "#A7CDF0", alpha = 0.5) +
  geom_line(color = "#4C72B0", linewidth = 0.8) +
  geom_point(color = "#4C72B0", size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  scale_x_continuous(breaks = cluster_candidates) +
  scale_y_continuous(breaks = cluster_candidates) +
  labs(
    x = "Requested clusters",
    y = "Non-empty clusters",
    title = "Cluster stability across repeated starts"
  ) +
  theme_bw()
```

## Cluster summaries

```{r mfuzz-cluster-summary}
cluster_summary |>
  mutate(
    MedianMaximumMembership = round(.data$MedianMaximumMembership, 3),
    EarlyMedianEffect = round(.data$EarlyMedianEffect, 3),
    LateMedianEffect = round(.data$LateMedianEffect, 3)
  ) |>
  knitr::kable(
    caption = paste0(
      "Mfuzz cluster summary. Core proteins have membership >= ",
      params$core_membership, "."
    )
  )
```

```{r mfuzz-standardized-profiles, fig.height=10, fig.cap="Standardized comparison-minus-reference protein effects used for Mfuzz clustering. Darker lines have higher membership in their assigned cluster; black lines are cluster centroids."}
standardized_profiles <- as.data.frame(Biobase::exprs(standardized_eset)) |>
  rownames_to_column("Protein") |>
  pivot_longer(
    cols = -Protein,
    names_to = "Time",
    values_to = "StandardizedEffect"
  ) |>
  mutate(Time = as.numeric(.data$Time)) |>
  inner_join(
    cluster_assignments |>
      select(.data$Protein, .data$Cluster, .data$MaximumMembership),
    by = "Protein"
  )

centroid_long <- centroid_table |>
  pivot_longer(
    cols = starts_with("Time"),
    names_to = "Time",
    values_to = "StandardizedEffect"
  ) |>
  mutate(Time = as.numeric(sub("^Time", "", .data$Time)))

ggplot(
  standardized_profiles,
  aes(
    x = .data$Time,
    y = .data$StandardizedEffect,
    group = .data$Protein,
    color = .data$MaximumMembership
  )
) +
  geom_line(alpha = 0.18, linewidth = 0.35) +
  geom_line(
    data = centroid_long,
    aes(
      x = .data$Time,
      y = .data$StandardizedEffect,
      group = .data$Cluster
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.2
  ) +
  geom_point(
    data = centroid_long,
    aes(
      x = .data$Time,
      y = .data$StandardizedEffect,
      group = .data$Cluster
    ),
    inherit.aes = FALSE,
    color = "black",
    size = 1.8
  ) +
  facet_wrap(~ Cluster, ncol = 2) +
  scale_color_gradient(low = "#D9EAF7", high = "#C44E52") +
  labs(
    x = "Time (hours)",
    y = paste("Standardized", effect_label, "effect"),
    color = "Membership",
    title = "Mfuzz differential-effect profiles"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r mfuzz-effects-log2, fig.height=9, fig.cap="Median comparison-minus-reference log2 effects and interquartile ranges within each assigned cluster. These are unstandardized effect sizes."}
effect_cluster_summary <- effect_long |>
  inner_join(
    cluster_assignments |>
      select(.data$Protein, .data$Cluster),
    by = "Protein"
  ) |>
  group_by(.data$Cluster, .data$Time) |>
  summarise(
    MedianEffect = stats::median(.data$Effect),
    LowerQuartile = stats::quantile(.data$Effect, 0.25),
    UpperQuartile = stats::quantile(.data$Effect, 0.75),
    .groups = "drop"
  )

ggplot(
  effect_cluster_summary,
  aes(x = .data$Time, y = .data$MedianEffect)
) +
  geom_hline(yintercept = 0, color = "grey55", linetype = "dotted") +
  geom_ribbon(
    aes(ymin = .data$LowerQuartile, ymax = .data$UpperQuartile),
    fill = "#A7CDF0",
    alpha = 0.55
  ) +
  geom_line(color = "#27445C", linewidth = 1) +
  geom_point(color = "#27445C", size = 2) +
  facet_wrap(~ Cluster, ncol = 2, scales = "free_y") +
  labs(
    x = "Time (hours)",
    y = paste(effect_label, "log2 abundance"),
    title = "Unstandardized comparison-effect trajectories"
  ) +
  theme_bw()
```

```{r mfuzz-group-means, fig.height=9, fig.cap="Median normalized protein abundances for the reference and comparison groups within each Mfuzz cluster. Ribbons show interquartile ranges across proteins."}
group_cluster_summary <- condition_summary |>
  inner_join(
    cluster_assignments |>
      select(.data$Protein, .data$Cluster),
    by = "Protein"
  ) |>
  group_by(.data$Cluster, .data$Time, .data$Group) |>
  summarise(
    MedianAbundance = stats::median(.data$MeanLog2Abundance),
    LowerQuartile = stats::quantile(.data$MeanLog2Abundance, 0.25),
    UpperQuartile = stats::quantile(.data$MeanLog2Abundance, 0.75),
    .groups = "drop"
  )

ggplot(
  group_cluster_summary,
  aes(
    x = .data$Time,
    y = .data$MedianAbundance,
    color = .data$Group,
    fill = .data$Group
  )
) +
  geom_ribbon(
    aes(ymin = .data$LowerQuartile, ymax = .data$UpperQuartile),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Cluster, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c("#4C72B0", "#C44E52")) +
  scale_fill_manual(values = c("#4C72B0", "#C44E52")) +
  labs(
    x = "Time (hours)",
    y = "Median normalized log2 abundance",
    color = NULL,
    fill = NULL,
    title = paste(params$reference_group, "and", params$comparison_group,
                  "profiles underlying each effect cluster")
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r mfuzz-core-table}
DT::datatable(
  cluster_assignments |>
    filter(.data$CoreMember) |>
    arrange(.data$Cluster, desc(.data$MaximumMembership)),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv")
  ),
  caption = paste0(
    "Mfuzz core proteins with maximum membership >= ",
    params$core_membership, "."
  )
)
```

# Output files

The complete tables are saved separately from the HTML report:

```{r output-files}
tibble(
  File = basename(c(
    filter_file,
    design_file,
    model_file,
    variable_file,
    condition_file,
    effect_file,
    membership_file,
    centroid_file,
    cluster_summary_file,
    diagnostic_file,
    file.path(output_dir, "timecourse_masigpro_mfuzz_objects.rds")
  )),
  Contents = c(
    "Coverage statistics for every protein",
    "Regression design submitted to maSigPro",
    "Overall and stepwise maSigPro model results",
    "Proteins selected for each maSigPro model term",
    "Observed group-time means and uncertainty",
    "Comparison-minus-reference effects used for clustering",
    "All Mfuzz memberships and assigned clusters",
    "Standardized Mfuzz cluster centroids",
    "Cluster sizes, core counts, and effect summaries",
    "Dmin and non-empty-cluster diagnostics",
    "Saved filtering, maSigPro, Mfuzz, and parameter objects"
  )
) |>
  knitr::kable(caption = "Machine-readable time-course analysis outputs.")
```

# Reproducibility

```{r session-info}
sessionInfo()
```
